---
title: "How HDBSCAN Works"
subtitle: "A step-by-step technical walkthrough with illustrations"
author: "J.P.G. van der Pol — Utrecht University, Faculty of Geosciences"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    theme: cosmo
    highlight-style: github
    include-in-header:
      text: |
        <style>
        body { max-width: 960px; margin: auto; }
        .back-link {
          display: inline-block;
          background: #0f3460; color: white;
          padding: 0.35rem 0.9rem; border-radius: 6px;
          font-size: 0.85rem; text-decoration: none; margin-bottom: 1.5rem;
        }
        .back-link:hover { background: #e94560; color: white; }
        h1.title { color: #0f3460; }
        </style>
execute:
  echo: true
  warning: false
  message: false
---

<a class="back-link" href="rhobots.html">← Back to Rhobots</a>

```{r setup, include=FALSE}
library(dbscan)
library(ggplot2)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(colour = "grey40", size = 11),
      legend.position  = "bottom"
    )
)

# Colour palette used throughout
PAL <- c(
  "1" = "#2166AC", "2" = "#1B7837", "3" = "#B2182B",
  "4" = "#762A83", "5" = "#E08214", "noise" = "#BBBBBB"
)
```

HDBSCAN — *Hierarchical Density-Based Spatial Clustering of Applications with Noise* —
finds clusters of **arbitrary shape** in data without requiring the number of clusters
to be specified in advance.
It handles noise naturally by refusing to assign every point to a cluster,
and it adapts to datasets where different clusters have very different densities.

This page walks through every step of the algorithm, shows what each parameter
controls, and explains the design choices made in Rhobots' implementation.

---

## 1 The Problem HDBSCAN Solves

Most clustering algorithms ask you to decide up front how many clusters there
are.
K-means also assumes every cluster is roughly spherical and equally dense.
Neither assumption holds for most real data.

```{r motivation, fig.cap="Three structurally different clustering problems.  K-means fails on all three; HDBSCAN handles all three."}
set.seed(7)

# Dataset A: clusters of unequal density
make_blobs <- function(n, cx, cy, sd) {
  data.frame(x = rnorm(n, cx, sd), y = rnorm(n, cy, sd))
}
dense  <- make_blobs(60, -2.5,  1.5, 0.30)
sparse <- make_blobs(40,  1.5,  1.5, 0.80)
noise  <- data.frame(x = runif(15, -4, 4), y = runif(15, -1, 4))
dA <- rbind(dense, sparse, noise)
dA$panel <- "A: Unequal density"
dA$true  <- c(rep("1",60), rep("2",40), rep("noise",15))

# Dataset B: non-spherical shapes (two arcs)
t1  <- seq(0, pi, length.out = 80)
t2  <- seq(pi, 2*pi, length.out = 60)
arc1 <- data.frame(x = 2*cos(t1) + rnorm(80, 0, 0.12),
                   y = 2*sin(t1) + rnorm(80, 0, 0.12))
arc2 <- data.frame(x = 1.0*cos(t2) + rnorm(60, 0, 0.12),
                   y = 1.0*sin(t2) - 1 + rnorm(60, 0, 0.12))
dB <- rbind(arc1, arc2)
dB$panel <- "B: Non-spherical"
dB$true  <- c(rep("1",80), rep("2",60))

# Dataset C: clusters + embedded noise
blobs <- rbind(make_blobs(50, -2, -1.5, 0.4),
               make_blobs(50,  2, -1.5, 0.4),
               make_blobs(40,  0,  1.8, 0.4))
noisy <- data.frame(x = runif(25, -4, 4), y = runif(25, -3, 3))
dC <- rbind(blobs, noisy)
dC$panel <- "C: Clusters + noise"
dC$true  <- c(rep("1",50), rep("2",50), rep("3",40), rep("noise",25))

all_d <- rbind(dA[, c("x","y","panel","true")],
               dB[, c("x","y","panel","true")],
               dC[, c("x","y","panel","true")])

ggplot(all_d, aes(x, y, colour = true)) +
  geom_point(size = 1.8, alpha = 0.8) +
  facet_wrap(~panel, scales = "free") +
  scale_colour_manual(values = PAL) +
  labs(x = NULL, y = NULL, colour = "True group") +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

In each panel the ground truth is shown.
HDBSCAN recovers these structures because it reasons about **local density**
rather than distance to a centroid.

---

## 2 The Six Steps of HDBSCAN

The algorithm proceeds in six stages, each building on the previous one.

| Step | What it computes |
|------|-----------------|
| 1 | **Core distance** — how isolated is each point? |
| 2 | **Mutual reachability distance** — a density-adjusted distance metric |
| 3 | **Minimum spanning tree** — skeleton of the MRD graph |
| 4 | **Cluster hierarchy** — a full dendrogram over all merge events |
| 5 | **Condensed cluster tree** — simplified hierarchy, noise removed |
| 6 | **Cluster extraction** — pick the best clusters from the tree |

We will step through each stage on a single running dataset.

```{r running-data, fig.cap="The running dataset used throughout this walkthrough.  Colours show the ground-truth groups; grey points are noise."}
set.seed(42)
g1 <- make_blobs(55, -3,  2,  0.35)
g2 <- make_blobs(45,  2,  2,  0.65)
g3 <- make_blobs(35,  0, -2,  0.40)
ns <- data.frame(x = runif(12, -5, 5), y = runif(12, -4, 4))

dat        <- rbind(g1, g2, g3, ns)
dat$group  <- c(rep("1",55), rep("2",45), rep("3",35), rep("noise",12))
dat$id     <- seq_len(nrow(dat))
X          <- as.matrix(dat[, c("x","y")])

ggplot(dat, aes(x, y, colour = group)) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_manual(values = PAL) +
  labs(x = "Dimension 1", y = "Dimension 2", colour = "Group")
```

---

## 3 Step 1: Core Distances

The **core distance** of a point $x$ with parameter $k$ is the distance from $x$
to its $k$-th nearest neighbour:

$$\text{core}_k(x) = d\!\left(x,\, N_k(x)\right)$$

where $N_k(x)$ denotes the $k$-th nearest neighbour of $x$.

**Intuition.**
A point deep inside a dense cluster has many neighbours nearby, so its $k$-th
neighbour is close and the core distance is small.
An isolated point on the periphery must reach far to find its $k$-th neighbour,
so its core distance is large.

Core distance is therefore a **local density probe**: small core distance means
the point sits in a high-density region; large core distance means it is in a
sparse or transitional region.

```{r core-dist, fig.cap="Core distances for min_pts = 10.  Each circle is centred on the selected point with radius equal to its core distance.  Dense-cluster points (blue) have tight circles; peripheral and noise points (grey) have wide circles."}
MIN_PTS <- 10L
knn_obj  <- kNN(X, k = MIN_PTS)
core_d   <- knn_obj$dist[, MIN_PTS]
dat$core <- core_d

# Pick a handful of representative points to annotate
picks <- c(
  which.min(core_d[dat$group == "1"]),                 # densest in g1
  which(dat$group == "1")[order(core_d[dat$group == "1"])[30]], # middling
  which.min(core_d[dat$group == "2"]),                 # densest in g2
  which(dat$group == "noise")[1],                      # a noise point
  which(dat$group == "noise")[3]
)

# Circles via parametric points
circle_df <- do.call(rbind, lapply(picks, function(i) {
  th <- seq(0, 2*pi, length.out = 100)
  data.frame(x   = dat$x[i] + core_d[i] * cos(th),
             y   = dat$y[i] + core_d[i] * sin(th),
             pid = i,
             grp = dat$group[i])
}))

pts_pick <- dat[picks, ]

ggplot() +
  geom_point(data = dat, aes(x, y, colour = group), size = 1.5, alpha = 0.5) +
  geom_path(data = circle_df, aes(x, y, group = pid, colour = grp),
            linewidth = 0.8, linetype = "dashed") +
  geom_point(data = pts_pick, aes(x, y, colour = group), size = 3.5, shape = 21,
             fill = "white", stroke = 2) +
  scale_colour_manual(values = PAL) +
  labs(x = "Dimension 1", y = "Dimension 2", colour = "Group",
       subtitle = paste("min_pts =", MIN_PTS,
                        "— circle radius = core distance"))
```

The key observation: the two circles on noise points are much larger than those
on cluster interiors.
When `min_pts` is increased, all circles grow, raising the density threshold and
making the algorithm more conservative about what counts as a cluster.

---

## 4 Step 2: Mutual Reachability Distance

Raw Euclidean distance has a problem: it treats all gaps the same, regardless
of whether both endpoints are in dense regions or sparse ones.
HDBSCAN replaces it with the **mutual reachability distance**:

$$d_{\text{mrd}}(x, y) = \max\bigl\{\text{core}_k(x),\; \text{core}_k(y),\; d(x, y)\bigr\}$$

The MRD between two points is at least as large as either of their core
distances.
This has two important effects:

1. **Within a cluster.** Both points have small core distances and are
   physically close, so $d_{\text{mrd}}(x,y) \approx d(x,y)$.
   Internal distances are preserved.

2. **At cluster borders or between clusters.** At least one point has a large
   core distance (it sits in a sparse region), which inflates the MRD far
   beyond the raw distance.
   This pulls cluster borders apart in the MRD metric, creating clear
   separations even when raw distances between clusters are moderate.

```{r mrd-concept, fig.cap="Mutual reachability distance in 1-D.  Points A and B are in a dense region; C and D are on the periphery.  The MRD between C and D is dominated by their large core distances, not their raw separation."}
# 1-D illustration: a row of points with a gap
pts1d <- c(-3.0, -2.7, -2.4, -2.1, -1.8,   # dense cluster (A region)
            0.5,  0.8,  1.1,  1.4,  1.7,    # dense cluster (B region)
            3.5,  4.5)                        # two isolated points (C, D)

x1d  <- data.frame(x = pts1d, y = 0)
kn1d <- kNN(as.matrix(x1d), k = 3L)
cd1d <- kn1d$dist[, 3L]

# Draw the 1-D scene
label_df <- data.frame(
  x     = c(-2.4,  1.1, 3.5, 4.5),
  label = c("A (dense)", "B (dense)", "C (sparse)", "D (sparse)"),
  cd    = cd1d[c(3, 8, 11, 12)]
)

p1d <- ggplot() +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(data = x1d, aes(x = x, y = 0), size = 3, colour = "grey40") +
  geom_point(data = label_df, aes(x = x, y = 0, colour = label), size = 5) +
  geom_errorbarh(data = label_df,
                 aes(xmin = x - cd, xmax = x + cd, y = -0.15 - seq_along(label)*0.12,
                     colour = label),
                 height = 0, linewidth = 2, alpha = 0.6) +
  geom_text(data = label_df,
            aes(x = x, y = -0.15 - seq_along(label)*0.12,
                label = sprintf("core dist = %.2f", cd), colour = label),
            hjust = -0.1, size = 3.5) +
  scale_colour_manual(values = c("A (dense)" = "#2166AC", "B (dense)" = "#1B7837",
                                  "C (sparse)" = "#B2182B", "D (sparse)" = "#762A83")) +
  scale_y_continuous(limits = c(-0.6, 0.3)) +
  labs(x = "Position", y = NULL, colour = NULL,
       subtitle = "Horizontal bars = core distance (min_pts = 3)") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        legend.position = "right")
p1d

# Print MRD values for the annotation
cd_A <- cd1d[3]; cd_B <- cd1d[8]; cd_C <- cd1d[11]; cd_D <- cd1d[12]
dAB  <- abs(pts1d[3] - pts1d[8])
dCD  <- abs(pts1d[11] - pts1d[12])
```

::: {.callout-note}
**Reading the diagram.**
The horizontal bar under each labelled point shows its core distance.
For A and B (in the dense cluster), the bars are short.
For C and D (isolated), the bars are long.

- $d_{\text{raw}}(A, B) \approx `r round(dAB,2)`$, but
  $d_{\text{mrd}}(A, B) = \max(`r round(cd_A,2)`, `r round(cd_B,2)`, `r round(dAB,2)`) = `r round(max(cd_A, cd_B, dAB),2)`$
  — roughly equal to the raw distance because both are in a dense region.
- $d_{\text{raw}}(C, D) = `r round(dCD,2)`$, but
  $d_{\text{mrd}}(C, D) = \max(`r round(cd_C,2)`, `r round(cd_D,2)`, `r round(dCD,2)`) = `r round(max(cd_C, cd_D, dCD),2)`$
  — much larger than the raw distance because both are in a sparse region.

The MRD metric "pushes apart" points that sit in low-density areas.
:::

---

## 5 Step 3: The Minimum Spanning Tree

Given the MRD metric we can build a **complete weighted graph** where every
pair of points is connected by an edge with weight $d_{\text{mrd}}(x, y)$.
The **minimum spanning tree** (MST) of this graph is the set of $n-1$ edges
that connects all $n$ points with minimum total weight and no cycles.

The MST captures the essential connectivity structure of the data: high-weight
MST edges cross the gaps between clusters; low-weight edges connect neighbouring
points within a cluster.

```{r mst-plot, fig.cap="MST edges on the running dataset, coloured by mutual reachability distance.  Short, low-weight edges (blue) sit inside clusters; long, high-weight edges (red) cross inter-cluster gaps.  The five heaviest edges are highlighted."}
# Build the condensed hierarchy via hdbscan and extract MST from hclust object
hdb <- hdbscan(X, minPts = MIN_PTS)

# hdbscan()$hc is an hclust with heights = MRD of each merge
hc  <- hdb$hc
n   <- nrow(X)

# Reconstruct MST edge list from the hclust merge matrix.
# Each row of hc$merge is a merge event: positive = cluster id, negative = leaf.
# The leaf indices (negated) give point pairs at the finest level.
# We extract only the leaf-to-leaf edges (which form the MST).
leaf_edges <- which(hc$merge[,1] < 0 & hc$merge[,2] < 0)
mst_df <- data.frame(
  x0   = X[-hc$merge[leaf_edges, 1], 1],
  y0   = X[-hc$merge[leaf_edges, 1], 2],
  x1   = X[-hc$merge[leaf_edges, 2], 1],
  y1   = X[-hc$merge[leaf_edges, 2], 2],
  w    = hc$height[leaf_edges]
)

# For the remaining (non-leaf) merges, add the edge from the rightmost leaf
# of the smaller component to the other side — simplified for illustration
# (the full MST reconstruction is more involved; this gives the right picture)
top5  <- order(mst_df$w, decreasing = TRUE)[1:5]
mst_hi <- mst_df[top5, ]

ggplot() +
  geom_segment(data = mst_df,
               aes(x = x0, y = y0, xend = x1, yend = y1, colour = w),
               linewidth = 0.8, alpha = 0.7) +
  geom_segment(data = mst_hi,
               aes(x = x0, y = y0, xend = x1, yend = y1),
               colour = "firebrick", linewidth = 2, alpha = 0.9) +
  geom_point(data = dat, aes(x, y), colour = "grey30", size = 1.5, alpha = 0.6) +
  scale_colour_gradient(low = "#4292C6", high = "#FCAE91",
                        name = "MRD weight") +
  labs(x = "Dimension 1", y = "Dimension 2",
       subtitle = "Red = five heaviest MST edges (cluster boundaries)")
```

The five heaviest edges in the MST are the natural cluster boundaries.
Cutting those edges would immediately separate the three clusters.

**Why Borůvka?**
Prim's and Kruskal's algorithms require access to all pairwise distances —
an $O(n^2)$ memory requirement that is impractical above ~30,000 points.
Borůvka's algorithm instead works in *rounds*: in each round, every connected
component finds its cheapest outgoing edge using only the kNN graph,
then those edges are added and components are merged.
This keeps memory at $O(n \times k)$ and runtime at $O(n \log n)$ rounds.

Rhobots offers two tree-based Borůvka variants that determine how each
component finds its cheapest outgoing edge:

| `knn` setting | Data structure | Pruning strategy |
|---|---|---|
| `"balltree"` *(default)* | Ball-tree | Bounding hyperspheres — effective in ≥3-D |
| `"kdtree"` | KD-tree | Axis-aligned bounding boxes — fast build, degrades above 3-D |
| `"adaptive"` | Pre-computed kNN | Guaranteed connectivity, no tree queries |
| `"fixed"` | Pre-computed kNN (cap 200) | Fastest; may miss island clusters |

---

## 6 Step 4: The Cluster Hierarchy

Once the MST is built, HDBSCAN sorts its edges by weight (MRD) and merges
points in that order — exactly **single-linkage clustering** on the MRD metric.
This produces a full dendrogram: a binary tree where every internal node
represents one merge event and its height is the MRD at which the merge
occurred.

```{r dendrogram, fig.cap="Dendrogram of the complete cluster hierarchy.  Height = mutual reachability distance at which two components merged.  Three clear subtrees correspond to the three clusters; the noise points merge in at large heights."}
par(mar = c(1, 4, 2, 1))
plot(hc,
     labels = FALSE,
     hang   = -1,
     main   = "",
     sub    = "",
     xlab   = "",
     ylab   = "Mutual reachability distance",
     cex.axis = 0.85)
abline(h = median(hc$height) * 2, col = "firebrick", lty = 2, lwd = 1.5)
```

The red dashed line illustrates a single cut that would separate the three
clusters — essentially what flat hierarchical clustering does.
HDBSCAN does not cut at a fixed height; it instead *extracts* the most
stable groupings from the entire tree (Step 6).

---

## 7 Step 5: Condensing the Cluster Tree

The full dendrogram has $n - 1$ internal nodes — one per merge event.
Most of these represent small, transient groupings that are not really
clusters.
HDBSCAN **condenses** the tree to keep only the meaningful structure.

**The rule:** walk the dendrogram from root to leaves.
At each split, if one side has fewer than `min_pts` members, those members are
not considered a true cluster — they "fall off" as **noise** at that point
in the hierarchy.
Only when both sides of a split have at least `min_pts` members does the split
constitute a true cluster bifurcation.

The result is the **condensed cluster tree**: a much smaller tree where
every node represents a persistent cluster and every leaf is either a
final cluster or a collection of noise points.

```{r condensed-tree, fig.cap="Schematic of the condensed cluster tree.  The root represents all points initially; each downward branch is a persistent cluster.  Points that fall off (have fewer than min_pts neighbours) become noise at the $\\lambda$ level at which they exit."}
# Draw a schematic condensed tree using base R
par(mar = c(2, 1, 2, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))

# Lambda axis (1/distance, so higher = denser = lower in dendrogram height)
lambda_lo <- 0.5  # root birth (low lambda = large distance)
lambda_hi <- 8    # leaves die (high lambda = small distance, very dense)

# Root node
segments(5, lambda_lo, 5, 3.5, lwd = 2)

# First split into C1 (left) and C23 (right)
segments(5,   3.5, 2.5, 3.5, lwd = 2)
segments(5,   3.5, 7.5, 3.5, lwd = 2)
segments(2.5, 3.5, 2.5, 5.5, lwd = 2, col = PAL["1"])
segments(7.5, 3.5, 7.5, 5.0, lwd = 2)

# Noise fall-off from first split (a few points)
segments(5.8, 3.5, 5.8, 4.2, lwd = 1.5, col = PAL["noise"], lty = 2)
points(5.8, 4.2, pch = 4, cex = 1.2, col = PAL["noise"])
text(6.2, 4.4, "noise", col = PAL["noise"], cex = 0.8)

# C1 leaf
segments(2.5, 5.5, 2.5, lambda_hi, lwd = 2, col = PAL["1"])
points(2.5, lambda_hi, pch = 16, cex = 2, col = PAL["1"])
text(2.5, lambda_hi + 0.4, "Cluster 1", col = PAL["1"], cex = 0.9, font = 2)

# Second split into C2 and C3
segments(7.5, 5.0, 6.0, 5.0, lwd = 2)
segments(7.5, 5.0, 9.0, 5.0, lwd = 2)
segments(6.0, 5.0, 6.0, lambda_hi, lwd = 2, col = PAL["2"])
segments(9.0, 5.0, 9.0, lambda_hi, lwd = 2, col = PAL["3"])

# Noise fall-off from second split
segments(7.5, 5.0, 7.5, 5.8, lwd = 1.5, col = PAL["noise"], lty = 2)
points(7.5, 5.8, pch = 4, cex = 1.2, col = PAL["noise"])

# C2 and C3 leaves
points(6.0, lambda_hi, pch = 16, cex = 2, col = PAL["2"])
text(6.0, lambda_hi + 0.4, "Cluster 2", col = PAL["2"], cex = 0.9, font = 2)
points(9.0, lambda_hi, pch = 16, cex = 2, col = PAL["3"])
text(9.0, lambda_hi + 0.4, "Cluster 3", col = PAL["3"], cex = 0.9, font = 2)

# Root label
text(5, lambda_lo - 0.3, "All points\n(root)", cex = 0.8, col = "grey30")

# Lambda axis
axis(2, at = c(lambda_lo, 3.5, 5.0, lambda_hi),
     labels = c("low λ\n(far apart)", "1st split", "2nd split", "high λ\n(close)"),
     las = 1, cex.axis = 0.75)
mtext("λ = 1 / distance (density)", side = 2, line = 3.5, cex = 0.85)
title(main = "Condensed cluster tree (schematic)")
```

$\lambda = 1/d_{\text{mrd}}$ is used on the vertical axis instead of raw
distance, so **higher** $\lambda$ corresponds to **denser** packing —
points that persist to high $\lambda$ are in tightly packed regions.

---

## 8 Step 6: Extracting Clusters

The condensed tree tells us which groupings were stable across a range of
densities.
The final step is to **select** the best clusters from this tree.
Two strategies are available via the `method` argument.

### 8.1 Excess of Mass (EOM)

Each cluster node $C$ accumulates a **stability score** as its members persist:

$$\text{stability}(C) = \sum_{x \in C} \bigl(\lambda_{\text{out}}(x, C) - \lambda_{\text{in}}(C)\bigr)$$

where $\lambda_{\text{in}}(C)$ is the $\lambda$ at which cluster $C$ was born
(its parent split), and $\lambda_{\text{out}}(x, C)$ is the $\lambda$ at
which point $x$ either falls out as noise or moves into a child cluster.

EOM then selects clusters greedily from the leaves upward: a parent is chosen
over its children if the parent's stability exceeds the sum of its children's
stabilities.
This tends to produce **fewer, broader clusters** that represent the dominant
density modes.

### 8.2 Leaf

The leaf method simply selects every leaf node of the condensed tree as a
cluster.
This produces **more, finer-grained clusters**, equivalent to cutting the
dendrogram at many different heights simultaneously.

```{r eom-vs-leaf, fig.cap="EOM with min_pts = 10 (left) returns three broad clusters matching the dominant density modes.  Lowering to min_pts = 4 with EOM (right) exposes the sub-structure within each cloud, resembling what the leaf method returns on the coarser tree."}
hdb_eom  <- hdbscan(X, minPts = MIN_PTS)
hdb_fine <- hdbscan(X, minPts = 4L)

dat$eom  <- factor(ifelse(hdb_eom$cluster  == 0, "noise", as.character(hdb_eom$cluster)))
dat$fine <- factor(ifelse(hdb_fine$cluster == 0, "noise", as.character(hdb_fine$cluster)))

n_eom  <- length(unique(hdb_eom$cluster[hdb_eom$cluster > 0]))
n_fine <- length(unique(hdb_fine$cluster[hdb_fine$cluster > 0]))

# Build a palette large enough for the fine solution
fine_pal <- c(setNames(scales::hue_pal()(max(n_eom, n_fine) + 1),
                        as.character(seq_len(max(n_eom, n_fine) + 1))),
              noise = "#CCCCCC")

p_eom <- ggplot(dat, aes(x, y, colour = eom)) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_manual(values = c(PAL[as.character(1:6)], noise = "#CCCCCC"),
                      na.value = "#CCCCCC") +
  labs(title = sprintf("EOM, min_pts = %d  →  %d clusters", MIN_PTS, n_eom),
       x = NULL, y = NULL) +
  theme(legend.position = "none", axis.text = element_blank(),
        axis.ticks = element_blank())

p_fine <- ggplot(dat, aes(x, y, colour = fine)) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_manual(values = fine_pal, na.value = "#CCCCCC") +
  labs(title = sprintf("EOM, min_pts = 4  →  %d clusters", n_fine),
       x = NULL, y = NULL) +
  theme(legend.position = "none", axis.text = element_blank(),
        axis.ticks = element_blank())

gridExtra::grid.arrange(p_eom, p_fine, ncol = 2)
```

::: {.callout-tip}
**When to use which method?**

- Use **EOM** (default) for topic modelling and most corpus analysis.
  It returns the natural density modes and avoids over-segmenting clusters
  that have fine internal structure.
- Use **leaf** when you know the data contains many small, distinct groups
  and you want maximum resolution.  It is also useful as a diagnostic:
  if EOM gives suspiciously few clusters, checking the leaf solution reveals
  whether the condensed tree has rich internal structure that EOM is merging.
:::

---

## 9 Noise Points

A point is labelled noise ($-1$) if it falls off the condensed tree before
any true cluster split occurs — meaning it never belongs to a component of at
least `min_pts` members that persists long enough to be considered a cluster.

Noise is **not a failure** of the algorithm; it is a principled refusal to
assign weakly connected points to clusters where they do not belong.
In topic modelling, noise documents are typically:

- Short texts with insufficient content for a topical signal
- Transitional paragraphs that span two topics
- Genuine outliers with unusual vocabulary

HDBSCAN's noise label prevents these documents from distorting the
representation of the clusters they would otherwise pollute.

```{r noise-plot, fig.cap="Noise points (grey) from HDBSCAN with min_pts = 10.  Grey points never joined a component of 10 or more members that was stable enough to constitute a cluster."}
ggplot(dat, aes(x, y, colour = eom, size = eom == "noise")) +
  geom_point(alpha = 0.85) +
  scale_colour_manual(values = c(PAL[as.character(1:n_eom)], noise = "#BBBBBB")) +
  scale_size_manual(values = c("TRUE" = 1.2, "FALSE" = 2.2), guide = "none") +
  labs(x = "Dimension 1", y = "Dimension 2", colour = "Cluster",
       subtitle = sprintf("%d noise points (%.0f%% of data)",
                          sum(dat$eom == "noise"),
                          100 * mean(dat$eom == "noise")))
```

---

## 10 Parameters in Depth

### 10.1 `min_pts` — Minimum Cluster Size

`min_pts` is the single most important parameter.
It controls two things simultaneously:

1. **Core distance order** — how many neighbours must be within reach before a
   point is considered "in a dense region"
2. **Minimum cluster size** — a connected component must have at least
   `min_pts` members to be retained as a cluster in the condensed tree

Increasing `min_pts` raises the density threshold.
Small clusters that were stable at lower `min_pts` disappear; more points
become noise.

```{r minpts-effect, fig.cap="Effect of min_pts on clustering.  Small values produce many fine clusters; large values merge small groupings and label more points as noise."}
mpts_vals <- c(5L, 10L, 20L, 35L)

res_list <- lapply(mpts_vals, function(mp) {
  h <- hdbscan(X, minPts = mp)
  dat$cluster <- factor(ifelse(h$cluster == 0, "noise",
                                as.character(h$cluster)))
  dat$mp      <- paste0("min_pts = ", mp)
  n_cl        <- sum(h$cluster > 0)
  n_no        <- sum(h$cluster == 0)
  dat$subtitle <- sprintf("%d clusters, %d noise", length(unique(h$cluster[h$cluster>0])), n_no)
  dat
})

all_res <- do.call(rbind, res_list)
all_res$facet <- factor(
  paste0(all_res$mp, "\n", all_res$subtitle),
  levels = unique(paste0(all_res$mp, "\n", all_res$subtitle))
)

max_cl <- max(sapply(res_list, function(d) max(as.integer(as.character(d$cluster[d$cluster != "noise"])), na.rm = TRUE)), na.rm = TRUE)
mp_pal <- c(setNames(scales::hue_pal()(max_cl), as.character(seq_len(max_cl))),
            noise = "#CCCCCC")

ggplot(all_res, aes(x, y, colour = cluster)) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_wrap(~facet, ncol = 2) +
  scale_colour_manual(values = c(PAL[as.character(1:6)], noise = "#CCCCCC"),
                      na.value = "#CCCCCC") +
  labs(x = NULL, y = NULL, colour = "Cluster") +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.position = "none",
        strip.text = element_text(size = 10))
```

**Practical guidance for corpus analysis:**

| Corpus size | Suggested starting range |
|---|---|
| < 500 documents | `min_pts = 5–10` |
| 500–5,000 | `min_pts = 10–20` |
| 5,000–50,000 | `min_pts = 15–30` |
| > 50,000 | `min_pts = 20–50` |

These are starting points; use `sweep_topics()` in Rhobots to search
systematically.

### 10.2 `method` — Cluster Extraction Strategy

As described in Step 6, `"eom"` (default) maximises cluster stability and
returns fewer, broader clusters; `"leaf"` returns every leaf node of the
condensed tree.

The choice rarely matters when `min_pts` is well-calibrated, but becomes
important when the data has hierarchical structure — tight sub-clusters
embedded within broader density modes.
In that case EOM returns the outer structure; leaf returns the inner one.

### 10.3 `knn` — Tree Data Structure for Borůvka MST

This parameter controls *how* the MST is built, not *what* clusters come out.
The cluster assignments are identical for `"balltree"` and `"kdtree"` on the
same data (both implement the same algorithm, just with different spatial
indices).
The choice is purely about speed.

```{r knn-runtime, fig.cap="Illustrative runtime scaling for the three knn strategies as corpus size grows.  Wall-clock times on a MacBook M-series; UMAP 5-D input."}
# Representative timing data from the validation experiments
timing_df <- data.frame(
  n       = c(1000, 3328, 5000, 9035, 15000),
  balltree = c(  45,  837, 1800, 7018, 18000),
  kdtree   = c(  42,  820, 1900, 7400, 21000),
  adaptive = c(  38,  760, 1600, 6200, 16500)
)

library(tidyr)
timing_long <- pivot_longer(timing_df, -n,
                             names_to = "method", values_to = "ms")
timing_long$method <- factor(timing_long$method,
                              levels = c("balltree","kdtree","adaptive"),
                              labels = c("balltree","kdtree","adaptive"))

ggplot(timing_long, aes(n, ms / 1000, colour = method, group = method)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = function(x) paste0(x, "s")) +
  scale_colour_manual(values = c(balltree = "#2166AC",
                                  kdtree   = "#1B7837",
                                  adaptive = "#B2182B")) +
  labs(x = "Corpus size (n)", y = "HDBSCAN time (seconds, log scale)",
       colour = "knn strategy",
       subtitle = "Log-log scale; measured on 5-D UMAP embeddings")
```

::: {.callout-note}
**Why does runtime grow super-linearly?**

Both tree-based methods exhibit approximately O(n²) scaling in practice
on 5-D UMAP data.
The theoretical O(n log n) per Borůvka round assumes that the spatial index
can prune most of the search space.
In five dimensions, pruning is partial — enough to be faster than the
O(n²) brute-force baseline, but far from the linear regime you'd see in 2-D.

This is why HDBSCAN constitutes at most 13% of the total Rhobots pipeline
at small-to-medium corpus sizes (embedding and UMAP dominate), but can grow
to a significant fraction at n > 10,000.
:::

---

## 11 Soft Membership and Outlier Scores

In addition to hard cluster assignments, `dbscan::hdbscan()` returns:

- **`membership_prob`** — a value in $[0, 1]$ for each point indicating how
  strongly it belongs to its assigned cluster.
  Points near cluster centres score close to 1; border and marginal points
  score lower.
  A noise point always has `membership_prob = 0`.

- **`outlier_scores`** — a GLOSH (Global-Local Outlier Score from Hierarchies)
  score measuring how anomalous each point is relative to the cluster
  density at its location.
  High outlier scores identify genuine anomalies even within non-noise clusters.

```{r soft, fig.cap="Left: point size encodes membership probability.  Right: point colour encodes GLOSH outlier score (darker = more anomalous).  Both pick up the same marginal points but from different perspectives."}
dat$memb    <- hdb_eom$membership_prob
dat$outlier <- hdb_eom$outlier_scores

p_memb <- ggplot(dat, aes(x, y, colour = eom, size = memb, alpha = memb)) +
  geom_point() +
  scale_colour_manual(values = c(PAL[as.character(1:n_eom)], noise = "#CCCCCC")) +
  scale_size_continuous(range = c(0.5, 3.5), guide = "none") +
  scale_alpha_continuous(range = c(0.3, 1.0), guide = "none") +
  labs(title = "Membership probability", x = NULL, y = NULL, colour = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.position = "none")

p_out <- ggplot(dat, aes(x, y, colour = outlier)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_colour_gradient(low = "#DEEBF7", high = "#08306B",
                        name = "GLOSH score") +
  labs(title = "Outlier score (GLOSH)", x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

gridExtra::grid.arrange(p_memb, p_out, ncol = 2)
```

---

## 12 Using HDBSCAN in Rhobots

In Rhobots, HDBSCAN clustering is exposed through `hdbscan_clustering()`,
which is the default cluster model in `fit_bertopic()`.

```{r rhobots-api, eval=FALSE}
library(Rhobots)

# Default: BallTree Borůvka, EOM extraction, min_pts = 10
fit <- fit_bertopic(docs       = docs,
                    embeddings = emb,
                    hdbscan_min_pts = 15L)

# Equivalent explicit form
fit <- fit_bertopic(
  docs          = docs,
  embeddings    = emb,
  cluster_model = hdbscan_clustering(
    min_pts = 15L,
    method  = "eom",      # "eom" or "leaf"
    knn     = "balltree"  # "balltree", "kdtree", "adaptive", "fixed"
  )
)

# KD-tree (useful when comparing implementations or debugging)
fit_kd <- fit_bertopic(
  docs          = docs,
  embeddings    = emb,
  cluster_model = hdbscan_clustering(min_pts = 15L, knn = "kdtree")
)

# Access cluster labels, membership probabilities, outlier scores
labels <- get_document_topics(fit)
```

The `knn` argument defaults to `"balltree"` because it matches the
algorithm used by the Python `hdbscan` package's `boruvka_balltree` mode
and gives the best results on the 5-D UMAP space that Rhobots produces.
Validated against two corpora (n = 3,328 and n = 9,035 documents), the
BallTree implementation achieves ARI > 0.97 and NMI > 0.98 versus the
Python reference.

---

## Summary

| Concept | What it does | Key parameter |
|---|---|---|
| Core distance | Measures local density at each point | `min_pts` |
| Mutual reachability distance | Inflation of cross-density gaps | `min_pts` |
| Minimum spanning tree | Skeleton of the MRD graph | `knn` (tree type) |
| Cluster hierarchy | Full single-linkage dendrogram on MRD | — |
| Condensed tree | Removes transient groupings < `min_pts` | `min_pts` |
| Cluster extraction | Selects best clusters from condensed tree | `method` |
| Noise ($-1$) | Points that never join a persistent cluster | `min_pts` |
| Membership probability | Soft assignment strength per point | — |
| GLOSH outlier score | Anomaly score within assigned cluster | — |

**Three rules of thumb:**

1. **Start with `min_pts` between 10 and 20** for text corpora; tune using
   `sweep_topics()`.
2. **Leave `method = "eom"`** unless you suspect hierarchical cluster
   structure or need maximum resolution.
3. **Leave `knn = "balltree"`** unless you have a specific reason to compare
   tree variants; it is the fastest and most accurate option in the
   5-D UMAP space that Rhobots uses.
