---
title: "How UMAP Works"
subtitle: "A step-by-step technical walkthrough with illustrations"
author: "J.P.G. van der Pol — Utrecht University, Faculty of Geosciences"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    theme: cosmo
    highlight-style: github
    include-in-header:
      text: |
        <style>
        body { max-width: 960px; margin: auto; }
        .back-link {
          display: inline-block;
          background: #0f3460; color: white;
          padding: 0.35rem 0.9rem; border-radius: 6px;
          font-size: 0.85rem; text-decoration: none; margin-bottom: 1.5rem;
        }
        .back-link:hover { background: #e94560; color: white; }
        h1.title { color: #0f3460; }
        </style>
execute:
  echo: true
  warning: false
  message: false
---

<a class="back-link" href="rhobots.html">← Back to Rhobots</a>

```{r setup, include=FALSE}
library(uwot)
library(ggplot2)
library(gridExtra)
library(scales)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor  = element_blank(),
      plot.title        = element_text(face = "bold", size = 14),
      plot.subtitle     = element_text(colour = "grey40", size = 11),
      axis.text         = element_blank(),
      axis.ticks        = element_blank(),
      legend.position   = "bottom"
    )
)

PAL <- c("1" = "#2166AC", "2" = "#1B7837", "3" = "#B2182B",
         "4" = "#762A83", "5" = "#E08214", "noise" = "#BBBBBB")
```

UMAP — *Uniform Manifold Approximation and Projection* — is a non-linear
dimensionality reduction algorithm that compresses high-dimensional data into
a low-dimensional space while preserving as much of the local and global
structure as possible.

In the Rhobots pipeline, UMAP sits between the embedding step and HDBSCAN:
384-dimensional sentence embeddings are first compressed to 5 dimensions by
UMAP, and HDBSCAN then finds clusters in that 5-D space.
The compression is necessary because HDBSCAN's distance computations become
unreliable above roughly 20 dimensions (the curse of dimensionality).

This page walks through every step of the UMAP algorithm, explains each
parameter, and describes the specific choices made in Rhobots.

---

## 1 Why Dimensionality Reduction?

In very high-dimensional spaces an uncomfortable thing happens: distances
between points converge.
A random point becomes roughly equidistant from all other points, which
means nearest-neighbour queries stop being informative — every point is
equally "close" to every other.

```{r curse, fig.cap="The curse of dimensionality.  As the number of dimensions grows, the ratio of maximum to minimum pairwise distance collapses toward 1 — all points become equidistant.  Nearest-neighbour structure dissolves."}
set.seed(1)
dims  <- c(2, 5, 10, 20, 50, 100, 200, 384)
ratios <- sapply(dims, function(d) {
  X  <- matrix(rnorm(500 * d), 500, d)
  D  <- as.matrix(dist(X))
  diag(D) <- NA
  mean(apply(D, 1, max, na.rm = TRUE) / apply(D, 1, min, na.rm = TRUE))
})

ggplot(data.frame(d = dims, r = ratios), aes(d, r)) +
  geom_line(colour = "#2166AC", linewidth = 1.2) +
  geom_point(colour = "#2166AC", size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 50, y = 1.08, label = "ratio = 1: all distances equal",
           colour = "grey50", size = 3.5) +
  scale_x_log10(breaks = dims, labels = dims) +
  labs(x = "Number of dimensions", y = "Mean (max dist / min dist) per point",
       subtitle = "500 random Gaussian points; ratio → 1 means k-NN becomes meaningless") +
  theme(axis.text = element_text(), axis.ticks = element_line())
```

At 384 dimensions (the output of `all-MiniLM-L6-v2`), the distance ratio is
already dangerously low.
UMAP addresses this by projecting into a space where nearest-neighbour
structure is preserved but distances are interpretable.

**Why not PCA?**
Principal Component Analysis finds the directions of maximum variance in the
data.
It is linear — it can only rotate and project, not bend or fold the data.
Structure that is non-linear (clusters connected through a curved manifold,
for instance) is not captured.

```{r pca-vs-umap, fig.cap="PCA (left) vs UMAP (right) on a 50-dimensional dataset containing three clusters embedded on non-linear 2-D manifolds.  PCA mixes the clusters; UMAP separates them."}
set.seed(42)
n_each <- 120

# Three clusters, each on a 2-D curved submanifold of 50-D space
make_cluster <- function(n, offset, seed) {
  set.seed(seed)
  t1   <- runif(n, -pi, pi)
  t2   <- runif(n, -1, 1)
  base <- cbind(t1, t2)                             # n x 2
  rot  <- qr.Q(qr(matrix(rnorm(50 * 2), 50, 2)))   # 50 x 2 orthonormal
  sweep(base %*% t(rot), 2, offset, "+") +           # n x 50
    matrix(rnorm(n * 50, 0, 0.15), n, 50)
}

Xhd <- rbind(
  make_cluster(n_each, c(3, 0, rep(0, 48)), 1),
  make_cluster(n_each, c(0, 3, rep(0, 48)), 2),
  make_cluster(n_each, c(0, 0, rep(0, 48)), 3)
)
labs_hd <- rep(c("1","2","3"), each = n_each)

# PCA to 2-D
pca_2d <- prcomp(Xhd, center = TRUE, scale. = FALSE)$x[, 1:2]

# UMAP to 2-D
umap_2d <- umap(Xhd, n_neighbors = 15, n_components = 2,
                min_dist = 0.1, metric = "euclidean",
                verbose = FALSE, seed = 42)

df_pca  <- data.frame(x = pca_2d[,1],  y = pca_2d[,2],  label = labs_hd, method = "PCA")
df_umap <- data.frame(x = umap_2d[,1], y = umap_2d[,2], label = labs_hd, method = "UMAP")
both    <- rbind(df_pca, df_umap)

ggplot(both, aes(x, y, colour = label)) +
  geom_point(size = 1.8, alpha = 0.75) +
  facet_wrap(~method, scales = "free") +
  scale_colour_manual(values = PAL) +
  labs(x = NULL, y = NULL, colour = "Cluster") +
  theme(axis.text = element_blank())
```

---

## 2 The Manifold Hypothesis

UMAP's mathematical foundation is the **manifold hypothesis**: data in
high-dimensional space does not fill that space uniformly — it lies on (or
near) a lower-dimensional **manifold** embedded within it.

A manifold is a surface that locally resembles flat Euclidean space but can
have global curvature.
The surface of a sphere is a 2-D manifold embedded in 3-D space.
The pages of a curled sheet of paper form a 2-D manifold in 3-D space.

```{r manifold-helix, fig.cap="A 1-D manifold (helix) embedded in 3-D space.  Each point is described by three coordinates, but the intrinsic structure is one-dimensional: position along the helix.  UMAP recovers this 1-D structure."}
set.seed(7)
n_h <- 400
t_h <- seq(0, 6 * pi, length.out = n_h)
helix <- data.frame(
  x = cos(t_h) + rnorm(n_h, 0, 0.06),
  y = sin(t_h) + rnorm(n_h, 0, 0.06),
  z = t_h / (2 * pi) + rnorm(n_h, 0, 0.04),
  t = t_h
)

# UMAP on 3-D helix → 1-D
h_umap <- umap(as.matrix(helix[, c("x","y","z")]),
               n_neighbors = 12, n_components = 1,
               min_dist = 0.01, metric = "euclidean",
               verbose = FALSE, seed = 42)

helix$umap1 <- h_umap[, 1]

p3d <- ggplot(helix, aes(x, y, colour = t)) +
  geom_path(aes(group = 1), colour = "grey80", linewidth = 0.3) +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_colour_viridis_c(option = "C", name = "Position\nalong helix") +
  labs(title = "3-D helix (data as seen)", x = "x", y = "y",
       subtitle = "Colour = true 1-D position") +
  theme(axis.text = element_text(), axis.ticks = element_line())

p1d <- ggplot(helix, aes(umap1, 0, colour = t)) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_viridis_c(option = "C", guide = "none") +
  labs(title = "UMAP to 1-D (recovered manifold)",
       x = "UMAP dimension 1", y = NULL,
       subtitle = "Intrinsic structure recovered") +
  scale_y_continuous(limits = c(-0.5, 0.5)) +
  theme(axis.text.y = element_blank())

grid.arrange(p3d, p1d, ncol = 1, heights = c(3, 1))
```

For sentence embeddings, the manifold hypothesis says: although documents are
represented as 384-dimensional vectors, the semantic structure of the corpus
occupies a much lower-dimensional manifold.
Documents about the same topic cluster together on this manifold, regardless
of their literal token overlap.

---

## 3 The Four Algorithm Steps

UMAP builds its embedding in four stages.

| Step | What it computes |
|------|-----------------|
| 1 | **k-nearest-neighbour graph** — who is close to whom in high-D? |
| 2 | **Fuzzy membership weights** — how strongly is each pair connected? |
| 3 | **Low-D initialisation** — a starting layout for optimisation |
| 4 | **Stochastic gradient descent** — push the low-D layout to match the high-D weights |

---

## 4 Step 1: The k-Nearest-Neighbour Graph

UMAP begins by finding the $k$ nearest neighbours of every point in
high-dimensional space using the chosen `metric`.
This produces a sparse graph where each point is connected to at most $k$
others.

The graph represents local structure: two points are connected if and only if
they are nearby in the original high-D space.
Global structure will emerge in Step 4 through the optimisation.

```{r knn-graph, fig.cap="k-nearest-neighbour graph on a small 2-D dataset (n_neighbors = 5).  Each point is connected to its 5 closest neighbours.  The graph is directed (arrows) before symmetrisation.  Dense regions form tight cliques; sparse regions have longer, weaker connections."}
set.seed(42)
n_small <- 60
Xs <- rbind(
  cbind(rnorm(25, -2, 0.5), rnorm(25, 1, 0.5)),
  cbind(rnorm(20,  2, 0.7), rnorm(20, 1, 0.7)),
  cbind(rnorm(15,  0, 0.4), rnorm(15, -2, 0.4))
)
lab_s <- c(rep("1",25), rep("2",20), rep("3",15))

K <- 5L
knn_s <- dbscan::kNN(Xs, k = K)

edge_df <- do.call(rbind, lapply(seq_len(n_small), function(i) {
  data.frame(x0 = Xs[i, 1], y0 = Xs[i, 2],
             x1 = Xs[knn_s$id[i, ], 1],
             y1 = Xs[knn_s$id[i, ], 2],
             w  = 1 / (knn_s$dist[i, ] + 1e-6))
}))

pts_df <- data.frame(x = Xs[,1], y = Xs[,2], label = lab_s)

ggplot() +
  geom_segment(data = edge_df, aes(x=x0, y=y0, xend=x1, yend=y1),
               colour = "grey60", linewidth = 0.4, alpha = 0.6,
               arrow = arrow(length = unit(0.08, "inches"), type = "open")) +
  geom_point(data = pts_df, aes(x, y, colour = label), size = 3, alpha = 0.9) +
  scale_colour_manual(values = PAL) +
  labs(x = NULL, y = NULL, colour = "Cluster",
       subtitle = paste("n_neighbors =", K)) +
  theme(axis.text = element_blank())
```

---

## 5 Step 2: Fuzzy Membership Weights

Raw distances are not directly usable because data density varies across the
space.
UMAP converts each edge distance into a **fuzzy membership weight** that
accounts for local density.

For point $x_i$, define:

$$\rho_i = \min_{j \in N_k(x_i)} d(x_i, x_j)$$

the distance to the *nearest* neighbour (not the $k$-th).
This anchors the local scale: a point deep in a dense cluster has a small
$\rho_i$; an isolated point has a large one.

The membership weight of the edge from $x_i$ to neighbour $x_j$ is then:

$$w_{ij} = \exp\!\left(-\frac{d(x_i, x_j) - \rho_i}{\sigma_i}\right)$$

where $\sigma_i$ is a per-point bandwidth calibrated so that the sum of
outgoing weights equals $\log_2(k)$.
Intuitively: the nearest neighbour always gets weight 1; further neighbours
get exponentially lower weights; the rate of decay is faster in dense regions
(small $\sigma_i$) and slower in sparse ones (large $\sigma_i$).

After computing directed weights $w_{ij}$ and $w_{ji}$ for each edge, UMAP
**symmetrises** the graph:

$$w^{\text{sym}}_{ij} = w_{ij} + w_{ji} - w_{ij} \cdot w_{ji}$$

This is the union of two fuzzy sets: if either direction is confident the
edge exists, it is retained.

```{r fuzzy-weights, fig.cap="Fuzzy membership weight as a function of distance from point $x_i$.  The curve shifts right in sparse regions (large $\\rho_i$, slow decay) and left in dense regions (small $\\rho_i$, fast decay).  This normalises for local density."}
sigma_dense  <- 0.3
sigma_sparse <- 1.2
rho_dense    <- 0.05
rho_sparse   <- 0.60

weight_fn <- function(d, rho, sigma) {
  ifelse(d <= rho, 1, exp(-(d - rho) / sigma))
}

d_seq <- seq(0, 4, length.out = 300)
wd <- data.frame(
  d      = rep(d_seq, 2),
  w      = c(weight_fn(d_seq, rho_dense,  sigma_dense),
             weight_fn(d_seq, rho_sparse, sigma_sparse)),
  region = rep(c("Dense region (small σ)", "Sparse region (large σ)"), each = 300)
)

ggplot(wd, aes(d, w, colour = region)) +
  geom_line(linewidth = 1.3) +
  geom_vline(xintercept = rho_dense,  linetype = "dashed",
             colour = PAL["1"], alpha = 0.5) +
  geom_vline(xintercept = rho_sparse, linetype = "dashed",
             colour = PAL["3"], alpha = 0.5) +
  annotate("text", x = rho_dense  + 0.05, y = 0.85,
           label = "ρ (dense)",  colour = PAL["1"], hjust = 0, size = 3.5) +
  annotate("text", x = rho_sparse + 0.05, y = 0.85,
           label = "ρ (sparse)", colour = PAL["3"], hjust = 0, size = 3.5) +
  scale_colour_manual(values = c("Dense region (small σ)"  = PAL["1"],
                                  "Sparse region (large σ)" = PAL["3"])) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = "Distance from xᵢ", y = "Membership weight w(xᵢ, xⱼ)",
       colour = NULL,
       subtitle = "Weight = 1 at the nearest neighbour; decays with distance") +
  theme(axis.text = element_text(), axis.ticks = element_line(),
        legend.position = "top")
```

---

## 6 Step 3: Initialising the Low-D Embedding

Before optimisation can start, every point needs an initial position in the
low-D target space.

The default initialisation (`init = "spectral"`) uses the eigenvectors of the
graph Laplacian of the symmetrised kNN graph.
This is the same computation as spectral clustering and gives a layout that
already reflects the large-scale connectivity — clusters are roughly in the
right neighbourhood even before any gradient descent.

An alternative is `init = "random"`, which places points uniformly at random
and relies entirely on the subsequent optimisation.
Spectral initialisation usually converges faster and produces more
reproducible results.

```{r init-comparison, fig.cap="Spectral initialisation (left) vs random initialisation (right) *before* any gradient descent, on the 3-cluster 50-D dataset.  Spectral already separates the clusters; random requires more SGD iterations to converge."}
Xl <- rbind(
  make_cluster(80, c(3, 0, rep(0,48)), 1),
  make_cluster(80, c(0, 3, rep(0,48)), 2),
  make_cluster(80, c(0, 0, rep(0,48)), 3)
)
labs_l <- rep(c("1","2","3"), each = 80)

init_spec <- umap(Xl, n_neighbors=15, n_components=2, n_epochs=0,
                  init="spectral", verbose=FALSE, seed=42)
init_rand <- umap(Xl, n_neighbors=15, n_components=2, n_epochs=0,
                  init="random",   verbose=FALSE, seed=42)

p_spec <- ggplot(data.frame(x=init_spec[,1], y=init_spec[,2], l=labs_l),
                 aes(x,y,colour=l)) +
  geom_point(size=1.8, alpha=0.8) +
  scale_colour_manual(values=PAL) +
  labs(title="Spectral init (n_epochs = 0)", x=NULL, y=NULL, colour=NULL) +
  theme(legend.position="none")

p_rand <- ggplot(data.frame(x=init_rand[,1], y=init_rand[,2], l=labs_l),
                 aes(x,y,colour=l)) +
  geom_point(size=1.8, alpha=0.8) +
  scale_colour_manual(values=PAL) +
  labs(title="Random init (n_epochs = 0)", x=NULL, y=NULL, colour=NULL) +
  theme(legend.position="none")

grid.arrange(p_spec, p_rand, ncol=2)
```

---

## 7 Step 4: Stochastic Gradient Descent

The final — and most expensive — step optimises the low-D positions to match
the high-D membership weights.

**Low-D membership.**
For two points $y_i$ and $y_j$ in the low-D embedding, the probability that
they are connected is modelled with a curve shaped by `min_dist` and `spread`:

$$q_{ij} = \Bigl(1 + a \,\|y_i - y_j\|^{2b}\Bigr)^{-1}$$

The parameters $a$ and $b$ are fitted from `min_dist` and `spread` so that
the curve falls to 0.5 at distance `min_dist`.
Small `min_dist` → steep curve → points are either very close or very far.
Large `min_dist` → shallow curve → points can sit at intermediate distances.

**Cross-entropy loss.**
UMAP minimises the cross-entropy between the high-D weights $w_{ij}$ and the
low-D probabilities $q_{ij}$:

$$\mathcal{L} = \sum_{(i,j) \in \text{edges}} \bigl[w_{ij} \log\tfrac{w_{ij}}{q_{ij}} + (1 - w_{ij}) \log\tfrac{1-w_{ij}}{1-q_{ij}}\bigr]$$

The first term is an **attractive force**: edges with high $w_{ij}$ (close
pairs in high-D) are pulled together in the low-D layout.
The second term is a **repulsive force**: pairs with low $w_{ij}$ (far apart
in high-D) are pushed apart.
UMAP uses *negative sampling* for the repulsive term — it samples random
non-edge pairs rather than iterating over all $O(n^2)$ pairs, which keeps
the per-iteration cost linear in $n$.

```{r low-d-curve, fig.cap="The low-D membership curve $q_{ij}$ for three values of min_dist.  Small min_dist (blue) creates a sharp threshold — points are either touching or repelled.  Large min_dist (red) allows a gradual spread."}
q_curve <- function(d, min_dist, spread = 1) {
  # Approximate a,b via a simple fit used in uwot
  b <- 0.8 + 0.15 * min_dist
  a <- 1 / (2 * min_dist^(2*b) * b)
  1 / (1 + a * d^(2*b))
}

d_seq2 <- seq(0, 3, length.out = 300)
md_vals <- c(0.0, 0.1, 0.5)
q_df <- do.call(rbind, lapply(md_vals, function(md) {
  data.frame(d = d_seq2,
             q = q_curve(d_seq2, max(md, 0.001)),
             md = paste0("min_dist = ", md))
}))
q_df$md <- factor(q_df$md, levels = paste0("min_dist = ", md_vals))

ggplot(q_df, aes(d, q, colour = md)) +
  geom_line(linewidth = 1.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 2.5, y = 0.52, label = "q = 0.5",
           colour = "grey50", size = 3.5) +
  scale_colour_manual(values = c("#2166AC","#1B7837","#B2182B")) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Distance in low-D space ||yᵢ − yⱼ||",
       y = "Low-D membership probability q(yᵢ, yⱼ)",
       colour = NULL,
       subtitle = "The curve falls to 0.5 at min_dist; controls how tightly clusters pack") +
  theme(axis.text = element_text(), axis.ticks = element_line(),
        legend.position = "top")
```

Gradient descent runs for `n_epochs` iterations (default 500 for small data,
200 for large).
The learning rate decays linearly over epochs so early iterations make large
moves to find the right neighbourhood structure and later iterations refine
local placement.

---

## 8 Parameters in Depth

### 8.1 `n_neighbors` — Local vs Global Structure

`n_neighbors` ($k$) is the most important UMAP parameter.
It controls how large a neighbourhood each point considers when building
the high-D graph.

- **Small $k$** (e.g. 5): each point only looks at its immediate neighbours.
  The embedding captures fine local structure; clusters may fragment and
  distant-but-related clusters may disconnect.
- **Large $k$** (e.g. 50): each point considers a wide neighbourhood.
  The embedding captures broad global structure; fine internal sub-clusters
  are smoothed away.

```{r n-neighbors, fig.cap="Effect of n_neighbors on the same 50-D dataset.  Small values expose fine structure; large values give a smoother, more globally coherent layout."}
set.seed(42)
Xn <- rbind(
  make_cluster(100, c(3, 0, rep(0,48)), 1),
  make_cluster(100, c(0, 3, rep(0,48)), 2),
  make_cluster(100, c(0, 0, rep(0,48)), 3)
)
labs_n <- rep(c("1","2","3"), each=100)

nn_vals <- c(5L, 15L, 30L, 50L)
nn_plots <- lapply(nn_vals, function(nn) {
  em <- umap(Xn, n_neighbors=nn, n_components=2, min_dist=0.1,
             metric="euclidean", verbose=FALSE, seed=42)
  df <- data.frame(x=em[,1], y=em[,2], label=labs_n)
  ggplot(df, aes(x,y,colour=label)) +
    geom_point(size=1.5, alpha=0.8) +
    scale_colour_manual(values=PAL) +
    labs(title=paste("n_neighbors =", nn), x=NULL, y=NULL) +
    theme(legend.position="none")
})
grid.arrange(grobs=nn_plots, ncol=2)
```

**Practical guidance for Rhobots:**
The default `n_neighbors = 15` works well for most corpora.
Increase toward 20–30 for very large corpora (n > 10,000) where you want
the manifold to reflect broad thematic similarity rather than narrow
document proximity.

### 8.2 `n_components` — Output Dimensionality

`n_components` controls the number of dimensions in the output space.

The two common choices in a BERTopic pipeline serve different purposes:

| `n_components` | Purpose |
|---|---|
| **2** | Visualisation — human-readable scatter plots |
| **5** *(Rhobots default)* | HDBSCAN input — more structure preserved, less distance distortion |
| 10–20 | Richer HDBSCAN input for very large or complex corpora |

Compressing all the way to 2-D for HDBSCAN input is tempting but problematic:
UMAP must throw away a lot of information to fit into two dimensions, and
the distortions can split coherent clusters or merge distinct ones.
Five dimensions preserves much more of the manifold structure while still
being far below 384.

```{r n-components, fig.cap="HDBSCAN on UMAP output at different n_components.  2-D (left) produces fragmented clusters; 5-D (right) recovers the correct three-cluster structure.  (Both scatter plots show a final 2-D UMAP for visualisation only.)"}
set.seed(42)
Xc <- rbind(
  make_cluster(90, c(3, 0, rep(0,48)), 1),
  make_cluster(90, c(0, 3, rep(0,48)), 2),
  make_cluster(90, c(0, 0, rep(0,48)), 3)
)
labs_c <- rep(c("1","2","3"), each=90)

umap2 <- umap(Xc, n_neighbors=15, n_components=2, min_dist=0.0, metric="euclidean", verbose=FALSE, seed=42)
umap5 <- umap(Xc, n_neighbors=15, n_components=5, min_dist=0.0, metric="euclidean", verbose=FALSE, seed=42)

hdb2 <- dbscan::hdbscan(umap2, minPts=8L)
hdb5 <- dbscan::hdbscan(umap5, minPts=8L)

# Visualise the 5-D result in 2-D via an additional UMAP pass
viz5 <- umap(umap5, n_neighbors=15, n_components=2, min_dist=0.1, verbose=FALSE, seed=42)

col_fn <- function(cl) ifelse(cl==0, "noise", as.character(cl))

p2 <- ggplot(data.frame(x=umap2[,1], y=umap2[,2],
                         cl=factor(col_fn(hdb2$cluster))),
             aes(x,y,colour=cl)) +
  geom_point(size=2, alpha=0.85) +
  scale_colour_manual(values=c(PAL[as.character(1:6)], noise="#CCCCCC"),
                      na.value="#CCCCCC") +
  labs(title=sprintf("n_components = 2  →  %d clusters",
                     length(unique(hdb2$cluster[hdb2$cluster>0]))),
       subtitle="(2-D UMAP used directly for HDBSCAN)",
       x=NULL, y=NULL) +
  theme(legend.position="none")

p5 <- ggplot(data.frame(x=viz5[,1], y=viz5[,2],
                         cl=factor(col_fn(hdb5$cluster))),
             aes(x,y,colour=cl)) +
  geom_point(size=2, alpha=0.85) +
  scale_colour_manual(values=c(PAL[as.character(1:6)], noise="#CCCCCC"),
                      na.value="#CCCCCC") +
  labs(title=sprintf("n_components = 5  →  %d clusters",
                     length(unique(hdb5$cluster[hdb5$cluster>0]))),
       subtitle="(5-D UMAP for HDBSCAN; 2-D UMAP for visualisation)",
       x=NULL, y=NULL) +
  theme(legend.position="none")

grid.arrange(p2, p5, ncol=2)
```

### 8.3 `min_dist` — How Tightly Points Cluster

`min_dist` sets the minimum distance between points in the low-D output.
It controls the `a` and `b` parameters of the $q_{ij}$ curve (Section 7).

- **`min_dist = 0.0`**: no minimum — points can be placed right on top of
  each other.
  Produces the tightest possible clusters.
  This is what Rhobots uses for the 5-D HDBSCAN input: you want points in
  the same topic to be as close as possible.
- **`min_dist = 0.1`** *(typical visualisation default)*: a small gap keeps
  individual points distinguishable in a scatter plot.
- **`min_dist > 0.5`**: points spread out; cluster structure softens.
  Useful when you care about global topology more than local clustering.

```{r min-dist, fig.cap="Effect of min_dist on the same dataset.  Low values (top left) compress clusters into tight blobs; high values (bottom right) spread the embedding out."}
set.seed(42)
Xm <- rbind(
  make_cluster(80, c(3, 0, rep(0,48)), 1),
  make_cluster(80, c(0, 3, rep(0,48)), 2),
  make_cluster(80, c(0, 0, rep(0,48)), 3)
)
labs_m <- rep(c("1","2","3"), each=80)

md_vals2 <- c(0.0, 0.1, 0.3, 0.8)
md_plots <- lapply(md_vals2, function(md) {
  em <- umap(Xm, n_neighbors=15, n_components=2,
             min_dist=md, metric="euclidean", verbose=FALSE, seed=42)
  df <- data.frame(x=em[,1], y=em[,2], label=labs_m)
  ggplot(df, aes(x,y,colour=label)) +
    geom_point(size=1.5, alpha=0.8) +
    scale_colour_manual(values=PAL) +
    labs(title=paste("min_dist =", md), x=NULL, y=NULL) +
    theme(legend.position="none")
})
grid.arrange(grobs=md_plots, ncol=2)
```

### 8.4 `metric` — Distance in High-D Space

The `metric` argument determines how distances between points are measured
*in the original high-D space* before any compression.

| Metric | Formula | Best for |
|---|---|---|
| `"euclidean"` | $\sqrt{\sum(x_i - y_i)^2}$ | Raw numeric features |
| `"cosine"` | $1 - \frac{x \cdot y}{\|x\|\|y\|}$ | Text embeddings (ignores magnitude) |
| `"manhattan"` | $\sum\|x_i - y_i\|$ | Sparse or count data |
| `"correlation"` | $1 - \text{Pearson}(x, y)$ | Time series, profiles |

**Why cosine for text embeddings?**
Sentence transformer models output vectors whose *direction* encodes meaning.
Two sentences about the same topic get vectors pointing in the same direction,
even if one sentence is short (small magnitude) and one is long (large
magnitude).
Cosine similarity ignores magnitude and only compares direction, so it
captures semantic similarity more reliably than Euclidean distance.

```{r metric, fig.cap="Cosine vs Euclidean metric on the same 50-D corpus-like data.  Cosine separates the three thematic clusters more cleanly because the cluster structure is directional, not based on absolute vector magnitudes."}
set.seed(42)
# Simulate embedding-like data: clusters differ in direction, not magnitude
make_embed_cluster <- function(n, direction, seed) {
  set.seed(seed)
  base  <- matrix(rnorm(n * 50, 0, 1), n, 50)
  dir_v <- direction / sqrt(sum(direction^2))
  sweep(base, 2, dir_v * 3, "+") *
    matrix(runif(n, 0.5, 2.0), n, 50)   # random magnitudes
}
d1 <- c(rep(1,10), rep(-1,40));  d1 <- d1/sqrt(sum(d1^2))
d2 <- c(rep(-1,10), rep(1,40)); d2 <- d2/sqrt(sum(d2^2))
d3 <- c(rep(1,25), rep(-1,25)); d3 <- d3/sqrt(sum(d3^2))

Xe <- rbind(make_embed_cluster(80, d1, 1),
            make_embed_cluster(80, d2, 2),
            make_embed_cluster(80, d3, 3))
labs_e <- rep(c("1","2","3"), each=80)

em_cos <- umap(Xe, n_neighbors=15, n_components=2, min_dist=0.1,
               metric="cosine",    verbose=FALSE, seed=42)
em_euc <- umap(Xe, n_neighbors=15, n_components=2, min_dist=0.1,
               metric="euclidean", verbose=FALSE, seed=42)

pe <- ggplot(data.frame(x=em_euc[,1], y=em_euc[,2], l=labs_e), aes(x,y,colour=l)) +
  geom_point(size=1.8, alpha=0.8) + scale_colour_manual(values=PAL) +
  labs(title='metric = "euclidean"', x=NULL, y=NULL) + theme(legend.position="none")
pc <- ggplot(data.frame(x=em_cos[,1], y=em_cos[,2], l=labs_e), aes(x,y,colour=l)) +
  geom_point(size=1.8, alpha=0.8) + scale_colour_manual(values=PAL) +
  labs(title='metric = "cosine"',    x=NULL, y=NULL) + theme(legend.position="none")
grid.arrange(pe, pc, ncol=2)
```

### 8.5 `seed` — Reproducibility

UMAP involves three stochastic components:

1. **Approximate nearest neighbours** (if using the default `nn_method = "annoy"`)
2. **Spectral initialisation** (uses random starting vectors for eigenvector solving)
3. **Negative sampling** during SGD

Setting `seed` passes the same value to all three sources.
Two runs with the same data, parameters, and seed will produce identical
embeddings up to floating-point precision.
*Without* a seed, the embedding topology is preserved but the overall rotation
and reflection may differ between runs — colours in cluster scatter plots
shift around but cluster shapes remain the same.

Rhobots passes `seed` through to `uwot::umap()` automatically when you
set `seed` in `fit_bertopic()`.

---

## 9 What UMAP Preserves (and What It Doesn't)

### Preserved ✓

- **Local neighbourhood structure**: if $x_i$ and $x_j$ are close in
  high-D, they will be close in the embedding.
- **Cluster separation**: well-separated clusters in high-D appear
  well-separated in the embedding.
- **Approximate topology**: the number of connected components and the broad
  shape of the manifold are maintained.

### Not Preserved ✗

- **Distances between clusters**: the absolute distances between clusters
  in the low-D embedding are not meaningful.
  Two clusters that appear far apart in the UMAP plot are not necessarily
  more different than two clusters that appear close — the optimisation only
  controls local structure reliably.
- **Cluster sizes**: a larger cluster in the embedding does not mean more
  documents; the number of points is the same regardless of visual area.
- **Density comparisons across clusters**: UMAP normalises local density
  (through $\sigma_i$), so a tight-looking cluster in the plot might not
  be denser than a spread-out cluster in the original space.

::: {.callout-warning}
**Inter-cluster distances in UMAP plots are not interpretable.**

It is tempting to conclude that clusters that are visually adjacent in a
2-D UMAP scatter are semantically more similar.
This is unreliable.
For meaningful cluster-to-cluster distance, compute the centroid distance in
the *original embedding space* (384-D), not in the UMAP projection.
:::

---

## 10 UMAP vs PCA

| | PCA | UMAP |
|---|---|---|
| Transformation | Linear | Non-linear |
| Preserves | Global variance | Local neighbourhood structure |
| Deterministic | Yes (up to sign) | No (stochastic; set seed) |
| Scales to large n | Yes ($O(nd^2)$) | Moderate ($O(n^{1.14})$ approx.) |
| Interpretable axes | Yes (principal components) | No |
| Handles manifold structure | No | Yes |
| Good for clustering input | Sometimes | Yes (with `min_dist = 0`) |
| Good for visualisation | Sometimes | Yes |

PCA remains the right choice when you need interpretable axes (factor
loadings, explained variance) or when you are reducing from a moderate
number of dimensions (e.g., 50 → 10) as a cheap preprocessing step.
UMAP is better when you need to preserve cluster structure for a downstream
clustering step or produce a human-readable scatter plot.

---

## 11 Using UMAP in Rhobots

In Rhobots, UMAP is called internally by `fit_bertopic()` and `sweep_topics()`.
The relevant arguments are:

```{r rhobots-api, eval=FALSE}
library(Rhobots)

# Default UMAP settings (n_neighbors=15, n_components=5, min_dist=0, cosine)
fit <- fit_bertopic(
  docs       = docs,
  embeddings = emb
)

# Adjust UMAP parameters
fit <- fit_bertopic(
  docs                  = docs,
  embeddings            = emb,
  umap_n_neighbors      = 20L,   # more global structure
  umap_n_components     = 10L,   # richer HDBSCAN input for large corpus
  umap_min_dist         = 0.0,   # tight clusters (always use 0 for HDBSCAN)
  seed                  = 42L    # reproducibility
)

# For visualisation only — 2-D UMAP after fitting
library(uwot)
viz <- umap(emb,
            n_neighbors = 15,
            n_components = 2,
            min_dist     = 0.1,       # spread out for readability
            metric       = "cosine",
            seed         = 42L)
```

::: {.callout-note}
**Two-stage UMAP in practice.**

Rhobots uses a single UMAP call to 5-D for clustering.
If you also want a 2-D visualisation, compute it as a *separate* UMAP
call from the original embeddings with `n_components = 2` and
`min_dist = 0.1`.
Do not project the 5-D output into 2-D — you would be reducing
already-reduced data and compounding distortions.
:::

### Why these defaults?

| Parameter | Rhobots default | Reason |
|---|---|---|
| `n_components` | 5 | Preserves more manifold structure than 2-D; well below 384-D curse-of-dimensionality threshold |
| `min_dist` | 0.0 | Maximises cluster tightness for HDBSCAN; no need to keep points visually separated |
| `metric` | `"cosine"` | Sentence embeddings encode meaning in direction, not magnitude |
| `n_neighbors` | 15 | Balances local detail and global coherence for typical corpus sizes |
| `seed` | user-supplied | Reproducibility of the full pipeline |

---

## Summary

| Concept | What it does | Key parameter |
|---|---|---|
| kNN graph | Encodes who is close to whom in high-D | `n_neighbors`, `metric` |
| Fuzzy membership weights | Converts distances to connectivity, normalised by local density | `n_neighbors` |
| Low-D curve $q_{ij}$ | Controls how tightly points pack in the output | `min_dist`, `spread` |
| SGD optimisation | Minimises cross-entropy between high-D weights and low-D probabilities | `n_epochs`, `learning_rate` |
| Output space | The compressed representation fed to HDBSCAN or plotted | `n_components` |

**Three rules of thumb for Rhobots:**

1. **Set `metric = "cosine"`** for all sentence-embedding inputs — it is
   the default and should stay that way unless you switch to a different
   encoder family.
2. **Use `min_dist = 0.0`** when the UMAP output will be passed to HDBSCAN;
   use `min_dist = 0.1`–`0.3` when producing scatter plots for humans.
3. **Increase `n_components` to 8–10** if `sweep_topics()` consistently
   finds fewer clusters than you expect — the 5-D projection may be
   discarding topic-discriminating structure in very large or heterogeneous
   corpora.
