---
title: "Rhobots: BERTopic-Style Topic Modeling in Pure R"
abstract: >
  We present \pkg{Rhobots}, an R package that implements the full BERTopic
  pipeline---transformer-based sentence embedding, UMAP dimensionality
  reduction, HDBSCAN clustering, and class-based TF-IDF topic
  extraction---without any dependency on Python, conda, or \pkg{reticulate}.
  Every stage runs natively in R through \pkg{torch}, \pkg{safetensors},
  \pkg{tok}, \pkg{uwot}, and \pkg{dbscan}.  The package exposes a modular,
  pluggable architecture in which dimensionality-reduction, clustering, and
  representation algorithms can be swapped independently.  Two representation
  models extend the default bag-of-words vocabulary: a POS-filtered model
  (via \CRANpkg{udpipe}) that restricts terms to selected Universal POS tags
  or multi-word POS-pattern phrases, and a C-value model
  \citep{frantzi2000cvalue} that selects n-grams based on their statistical
  evidence as domain-specific multi-word terms.  Beyond the core pipeline the
  package provides seven advanced analysis tools: zero-shot topic
  classification, topic stability analysis, lexical coherence assessment,
  guided topic modeling, LLM-based topic labeling, structured model
  persistence, and class-conditional group comparison.  We describe the
  design, demonstrate the package on a corpus of public-domain books sourced
  via Project Gutenberg, and compare the approach to established R topic
  models such as Latent Dirichlet Allocation and the Structural Topic Model.
author:
  - name: J.P.G. van der Pol
    affiliation: Utrecht University
    address: |
      Faculty of Geosciences, Vening Meineszgebouw A
      Princetonlaan 8a, 3584 CB Utrecht, The Netherlands
    url: https://www.uu.nl
    email: j.p.g.vanderpol@uu.nl
    orcid: 0000-0002-9686-9637
date: "2026-06-25"
date_received: ~
bibliography: rhobots.bib
#output: rjtools::rjournal_web_article
output: rjtools::rjournal_pdf_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
library(Rhobots)
```

# Introduction

Topic modeling occupies a central place in computational social science,
scientometrics, and digital humanities.  Classical approaches such as Latent
Dirichlet Allocation \citep[LDA;][]{blei2003lda} and its structural extension
\citep[STM;][]{roberts2019stm} are well established in R and remain the
default choice for many practitioners.  However, these bag-of-words models
discard word order and cannot capture the semantic similarity between
synonymous or paraphrastic expressions, a limitation that becomes increasingly
costly as corpora grow more heterogeneous.

The Python package BERTopic \citep{grootendorst2022bertopic} proposed an
alternative: embed every document into a dense contextual vector space using a
pre-trained sentence transformer, reduce the dimensionality of those embeddings
with UMAP \citep{mcinnes2018umap}, identify coherent clusters with HDBSCAN
\citep{campello2013hdbscan}, and finally label each cluster through a
class-based variant of TF-IDF (c-TF-IDF).  The result is a topic model whose
topics are semantically coherent, whose number of topics is determined
automatically by the data, and whose terms are directly interpretable.
BERTopic has attracted considerable attention and has become one of the most
widely used topic models in recent NLP and social-science work.

For R users, accessing this pipeline has historically required a working
Python installation, a compatible conda or virtual environment, and the
\CRANpkg{reticulate} bridge.  That dependency chain is version-sensitive,
difficult to maintain across operating systems, and a genuine barrier in
teaching and reproducibility settings.

\pkg{Rhobots} removes that barrier entirely.  It is a self-contained R
package that implements the complete BERTopic pipeline using only CRAN
packages:
\CRANpkg{torch} for transformer forward passes,
\CRANpkg{safetensors} for safe weight loading,
\CRANpkg{hfhub} for HuggingFace Hub model discovery,
\CRANpkg{tok} for fast WordPiece tokenization,
\CRANpkg{uwot} for UMAP,
\CRANpkg{dbscan} for HDBSCAN, and
\CRANpkg{Matrix} for sparse document-term matrices.
No Python installation is required or consulted at any point.

The package is designed around three goals.  First, *accessibility*: a topic
model should be fittable in a single function call from any standard R
installation.  Second, *modularity*: every stage of the pipeline should be
swappable so that practitioners can substitute PCA for UMAP, k-means for
HDBSCAN, or a domain-specific encoder for the default without rewriting the
pipeline.  Third, *API parity*: the accessor functions mirror the Python
BERTopic API so that users familiar with the Python package can transfer
knowledge directly.

The rest of this paper is organised as follows.  Section \@ref(sec:algorithm)
describes the BERTopic algorithm and the specific choices made in Rhobots.
Section \@ref(sec:design) details the package design and S3 class
architecture.  Section \@ref(sec:usage) documents all user-facing functions,
including the pluggable representation models and seven advanced analysis
tools.  Section \@ref(sec:example) presents a worked example on a corpus of
scientific abstracts.  Section \@ref(sec:validation) validates the R
implementation against the Python \texttt{sentence-transformers} reference.
Section \@ref(sec:casestudy) applies the full pipeline to a corpus of
public-domain books as a realistic end-to-end case study.
Section \@ref(sec:discussion) compares Rhobots to existing R topic models
and outlines limitations.  Section \@ref(sec:conclusion) concludes.

# The BERTopic Pipeline {#sec:algorithm}

BERTopic decomposes topic discovery into four sequential stages.  Figure
\@ref(fig:pipeline) provides a schematic overview.

## Stage 1: Sentence Embedding

Each document $d_i$ in the corpus is mapped to a dense vector
$\mathbf{e}_i \in \mathbb{R}^H$ by a pre-trained transformer encoder.
\pkg{Rhobots} supports any BERT-family model \citep{devlin2019bert} hosted
on the HuggingFace Hub, including domain-adapted variants such as SciBERT
\citep{beltagy2019scibert} and BERT-for-Patents.  The encoder performs a
standard forward pass: the input token sequence is embedded, passed through
$L$ transformer layers of multi-head self-attention and feed-forward
sublayers, and the final hidden states are mean-pooled over non-padding
positions to produce a single vector per document.  Vectors are L2-normalised
by default so that cosine similarity equals the dot product.

## Stage 2: Dimensionality Reduction

Raw BERT embeddings typically have $H = 384$ to $H = 768$ dimensions.
Clustering in such high-dimensional spaces suffers from the curse of
dimensionality: distance functions lose discriminative power and density
estimation becomes unreliable.  UMAP \citep{mcinnes2018umap} addresses this by
learning a low-dimensional manifold representation that preserves the local
neighbourhood structure of the original embedding space.  \pkg{Rhobots}
reduces to $k = 5$ components by default before clustering.  An independent
2-D projection is always computed for visualization.

The dimensionality-reduction step is pluggable: users can substitute PCA
(`pca_reduction()`), skip reduction altogether (`no_reduction()`), or pass any
object that implements the `dim_reduce` and `dim_project` S3 generics.

## Stage 3: Density-Based Clustering

HDBSCAN \citep{campello2013hdbscan} identifies a variable number of
variable-density clusters in the reduced space and assigns a noise label
($-1$) to documents that do not belong to any coherent cluster.  This is
important: unlike k-means or agglomerative clustering, HDBSCAN does not force
every document into a topic.  Documents that are genuinely off-topic---short
boilerplate, foreign-language text, or truly singular documents---are marked
as noise rather than distorting topic representations.

The clustering step is also pluggable: `kmeans_clustering()` and
`agglomerative_clustering()` are provided for settings where the number of
clusters is known in advance or where noise-free assignment is required.

## Stage 4: Class-Based TF-IDF

Once cluster assignments are known, \pkg{Rhobots} represents each topic
using a class-based variant of TF-IDF.  All documents assigned to cluster
$c$ are concatenated into one pseudo-document.  Within-class term frequency
(class-TF) is computed as the proportion of each term within that
pseudo-document, and class-IDF weights each term by the log of the mean
class size divided by the number of classes in which the term appears:

$$
\text{c-TF-IDF}_{t,c} = \text{TF}_{t,c} \times \log\!\left(1 + \frac{A}{f_t}\right),
$$

where $A$ is the mean number of words per class and $f_t$ is the number of
classes in which term $t$ appears at least once.  The top-$n$ terms by
c-TF-IDF score form the topic representation.  An optional
`reduce_frequent_words` flag applies a square-root transform to the class-TF
before multiplication, downweighting terms that are overwhelmingly dominant
within a single class---a direct port of the same option in the Python
BERTopic `ClassTfidfTransformer`.

```{r pipeline, echo=FALSE, fig.cap="The four-stage BERTopic pipeline as implemented in Rhobots. Each stage is independently pluggable through S3 generics.", fig.alt="Schematic showing four boxes: Embed (torch), Reduce (uwot/UMAP), Cluster (dbscan/HDBSCAN), and Represent (c-TF-IDF), connected by arrows."}
# Diagram is rendered as a simple flow illustration via base R graphics
par(mar = c(1, 1, 1, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 2))

boxes <- data.frame(
  x     = c(1, 3.5, 6, 8.5),
  label = c("Embed\n(torch)", "Reduce\n(uwot)", "Cluster\n(dbscan)", "Represent\n(c-TF-IDF)"),
  col   = c("#D6EAF8", "#D5F5E3", "#FAD7A0", "#F9EBEA")
)

for (i in seq_len(nrow(boxes))) {
  rect(boxes$x[i] - 0.9, 0.4, boxes$x[i] + 0.9, 1.6,
       col = boxes$col[i], border = "grey40", lwd = 1.5)
  text(boxes$x[i], 1.0, boxes$label[i], cex = 0.9, font = 2)
}

for (i in 1:3)
  arrows(boxes$x[i] + 0.9, 1.0, boxes$x[i + 1] - 0.9, 1.0,
         length = 0.12, lwd = 1.5, col = "grey30")
```

# Package Design {#sec:design}

## S3 Class Architecture

\pkg{Rhobots} defines three main S3 classes.

**`bert_encoder`** — returned by `load_hf_bert()`, wraps the `nn_module`,
tokenizer, and config.  The `print` method reports model size and vocabulary.

**`bertopic_fit`** — returned by `fit_bertopic()`, contains the full fitted
state: embeddings, reduced coordinates, 2-D layout, cluster assignments,
topic-term table, topic sizes, topic labels, representative documents, the
fitted dimensionality-reduction model, and the fitted clustering model.
Print, accessor, and manipulation methods dispatch on this class.

**`bertopic_flow`** — returned by `fit_topics_over_time()`, represents a
sequence of independently fitted topic models aligned by centroid similarity,
supporting the `visualize_topic_flow()` Sankey diagram.

## Pluggable Pipeline Models

Three extension points form the backbone of the pipeline.  The first two are
S3 generics:

```{r generics, eval=FALSE}
dim_reduce(model, X, seed, verbose)  # returns list(embedding, model)
dim_project(model, X)                # transforms new data into fitted space

cluster_docs(model, X, seed)         # returns list(labels, model)
```

Any object implementing these generics can be passed to `fit_bertopic()` via
the `dim_reduction_model` and `cluster_model` arguments, allowing new
reduction or clustering algorithms to be added without modifying the core
pipeline code.

The third extension point is `representation_model`.  This argument controls
how the document-term matrix is built for the c-TF-IDF step.  Two constructors
are provided: `pos_representation()` restricts the vocabulary to tokens
matching specified Universal POS tags (using \CRANpkg{udpipe} for annotation)
and optionally extracts multi-word phrases matching POS-tag sequences;
`cvalue_representation()` selects n-grams using the C-value method
\citep{frantzi2000cvalue}, which scores candidate multi-word terms by their
frequency discounted by how often they appear as part of a longer term.
When `representation_model = NULL` (the default) the existing
document-frequency-filtered bag-of-words vocabulary is used unchanged,
preserving full backward compatibility.

## HuggingFace Integration

`load_hf_bert()` downloads model weights, configuration, and tokenizer from
the HuggingFace Hub through \CRANpkg{hfhub}.  It tries to use
`model.safetensors` when available (safe, device-agnostic, fast) and falls
back to `pytorch_model.bin` when it is not.  It tries to use the fast
`tokenizer.json` via \CRANpkg{tok} and falls back to a WordPiece tokenizer
built from `vocab.txt` via \CRANpkg{wordpiece}.  A `weights_path` argument
allows pinning to a local file, which is useful when a model is available only
in a CUDA-saved `.bin` format that requires prior conversion.

## Embedding Cache

`embed_texts_cached()` wraps `embed_texts()` with a simple on-disk cache.
The first call saves the embedding matrix to an `.rds` file; subsequent calls
load from disk without running the encoder.  This is important in practice
because transformer inference over a large corpus can take several minutes on
a CPU, and it is desirable to separate the embedding step from the modeling
step in an analysis workflow.

# Package Functions {#sec:usage}

\pkg{Rhobots} organises its functions into nine groups: loading transformer
models, producing embeddings, fitting topic models, representation models,
inspecting results, post-fit manipulation, visualization, dynamic topic
modeling, and seven advanced analysis tools.  This section describes each
function, the problem it solves, and why it is important for a practical
topic-modeling workflow.

## Loading Transformer Models

### `load_hf_bert()`

`load_hf_bert(repo_id, weights_path = NULL)` is the primary entry point to
the transformer ecosystem in \pkg{Rhobots}.  It downloads and initialises a
BERT-family encoder directly from the HuggingFace Hub without any Python
runtime.  The function resolves three components automatically: the model
architecture from `config.json`, the weight file (preferring the safe and
device-agnostic `model.safetensors` format over the legacy `pytorch_model.bin`
where available), and the tokenizer (preferring the fast HuggingFace
`tokenizer.json` format via \CRANpkg{tok} and falling back to a WordPiece
tokenizer built from `vocab.txt`).

The `weights_path` argument allows pinning to a local weight file, which is
needed when a model's upstream repository provides only a CUDA-saved binary
that must be converted to safetensors before use.

`load_hf_bert()` is important because it replaces the entire Python-side model
loading infrastructure (HuggingFace `transformers`, `torch`, `tokenizers`) with
a single R function call.  Any BERT-family model --- BERT-base, MiniLM, MPNet,
SciBERT, BioBERT, BERT-for-Patents, FinBERT, and the sentence-transformer
variants built on all of these --- can be loaded this way.  The returned
encoder object is passed unchanged to all downstream embedding functions.

```{r load-hf, eval=FALSE}
enc <- load_hf_bert("sentence-transformers/all-MiniLM-L6-v2")
enc <- load_hf_bert("allenai/scibert_scivocab_uncased")
enc <- load_hf_bert("anferico/bert-for-patents",
                    weights_path = "/local/model.safetensors")
```

### `make_wordpiece_tokenizer()`

`make_wordpiece_tokenizer(vocab_path)` constructs a WordPiece tokenizer from a
raw `vocab.txt` vocabulary file via \CRANpkg{wordpiece}.  It exposes the same
`encode_batch()`, `enable_padding()`, and `enable_truncation()` interface as
`tok::tokenizer`, so `embed_texts()` cannot tell the difference.

This function matters for older BERT checkpoints --- SciBERT, BioBERT, the
original BERT-base releases --- that predate the HuggingFace fast-tokenizer
format.  `load_hf_bert()` calls it automatically when `tokenizer.json` is
absent, but it is exported so users who want to construct a custom tokenizer
from a locally modified vocabulary can do so directly.

## Producing Sentence Embeddings

### `embed_texts()`

`embed_texts(encoder, texts, batch_size, max_length, normalize, device)` is
the core embedding function.  It tokenizes the input texts, splits them into
batches, runs each batch through the transformer's forward pass, mean-pools
the final hidden states over non-padding token positions to obtain one vector
per document, and optionally L2-normalises each row so that cosine similarity
equals the dot product.

Batching is important because transformers operate on fixed-width tensors:
all sequences in a batch must be padded to the same length.  Batching in
memory-efficient chunks also prevents out-of-memory errors on large corpora.
The `device` argument enables GPU acceleration when \CRANpkg{torch} is built
with CUDA support.

### `mean_pool()`

`mean_pool(hidden, attention_mask)` is the low-level pooling primitive: given
a 3-D tensor of shape (batch $\times$ sequence $\times$ hidden) and the
corresponding attention mask, it zeroes out padding positions and takes the
mean across the sequence dimension.  It is exported so that users who want to
implement custom pooling strategies --- CLS-token pooling, max pooling, or
weighted pooling --- can call it directly inside their own forward pass.

### `embed_texts_cached()`

`embed_texts_cached(encoder, texts, cache_file, overwrite)` wraps
`embed_texts()` with on-disk caching.  On the first call it computes
embeddings and writes them to a `.rds` file.  On every subsequent call with
the same `cache_file` path it reads the matrix from disk and skips the encoder
entirely, after verifying that the cached row count matches the length of
`texts`.

This function is essential for any practical workflow.  Embedding a corpus of
several thousand documents with a medium-sized transformer takes minutes on
CPU.  Without caching, every change to clustering hyperparameters, stopword
lists, or ngram ranges would require re-running the encoder.  The cache
decouples the expensive embedding step from the cheap modeling step, making
iterative experimentation fast.

```{r cache-ex, eval=FALSE}
# First call: runs the encoder and writes embeddings.rds
emb <- embed_texts_cached(enc, docs, cache_file = "embeddings.rds")

# Subsequent calls: loads instantly, no encoder needed
emb <- embed_texts_cached(cache_file = "embeddings.rds", texts = docs)

# Invalidate the cache when docs change
emb <- embed_texts_cached(enc, new_docs,
                           cache_file = "embeddings.rds", overwrite = TRUE)
```

### `save_embeddings()` and `load_embeddings()`

`save_embeddings(embeddings, path)` and `load_embeddings(path)` are lower-level
persistence helpers.  They support both `.rds` (recommended: lossless, fast,
R-native) and `.csv` (portable to other tools).  These functions are useful
when sharing an embedding matrix between collaborators or between R sessions
without relying on `embed_texts_cached()`'s one-to-one file coupling.

## Fitting the Topic Model

### `fit_bertopic()`

`fit_bertopic()` is the main pipeline function.  It accepts the corpus, an
optional encoder or pre-computed embedding matrix, and model objects for
dimensionality reduction and clustering.  It executes the four-stage pipeline
described in Section \@ref(sec:algorithm) and returns a `bertopic_fit` object
containing the full fitted state.

Key arguments control every aspect of the fit.  `dim_reduction_model`,
`cluster_model`, and `representation_model` accept pluggable model objects
(see Sections \@ref(sec:design) and Representation Models below); legacy
keyword arguments (`umap_n_neighbors`, `hdbscan_min_pts`, etc.) are provided
for convenience and build the default objects automatically.  `ngram_range`
controls whether unigrams only (`c(1L, 1L)`) or unigrams and bigrams
(`c(1L, 2L)`) are included in the default bag-of-words DTM (ignored when a
`representation_model` is supplied).  `reduce_frequent_words` applies a
square-root dampening to the class-TF before c-TF-IDF scoring, which improves
interpretability when a small number of terms dominates a topic's vocabulary.

```{r fit-ex, eval=FALSE}
fit <- fit_bertopic(
  docs                  = docs,
  embeddings            = emb,
  hdbscan_min_pts       = 5,
  ngram_range           = c(1L, 2L),
  extra_stopwords       = c("study", "paper", "result"),
  reduce_frequent_words = TRUE
)
```

### `build_dtm()` and `c_tf_idf()`

These two functions form the term-extraction layer of the pipeline and are
exported for users who want to reproduce or customise the representation step.

`build_dtm(docs, min_df, max_df_frac, stopwords, ngram_range)` tokenizes a
character vector (lowercase, non-alphanumeric characters stripped), removes
stopwords, constructs n-grams, and returns a sparse `dgCMatrix` with one row
per document and one column per vocabulary term.  The `min_df` and
`max_df_frac` filters remove terms that are either too rare to be
interpretable or so ubiquitous that they carry no discriminative signal.

`c_tf_idf(dtm, cluster_ids, top_n, reduce_frequent_words)` takes the sparse
DTM and a vector of cluster assignments and computes the class-based TF-IDF
scores described in Section \@ref(sec:algorithm), returning a tidy data frame
of the top-$n$ terms per topic.

## Pluggable Pipeline Models

\pkg{Rhobots} provides three dimensionality-reduction and three clustering
constructors.  Each is a plain S3 object that can be passed to `fit_bertopic()`
via `dim_reduction_model` or `cluster_model`.

**`umap_reduction(n_neighbors, n_components, min_dist, metric)`** is the
default.  UMAP preserves local neighbourhood structure in the embedding space,
which is crucial for density-based clustering: if the manifold is not
well-separated in the reduced space, HDBSCAN cannot form meaningful clusters.
The `n_components` parameter controls the dimensionality of the clustering
space (default 5); a separate 2-D projection is always computed by
`fit_bertopic()` for visualization.

**`pca_reduction(n_components, scale.)`** wraps `stats::prcomp`.  PCA is
fully deterministic, substantially faster than UMAP for large corpora, and
avoids the stochastic variation that UMAP introduces.  The trade-off is that
PCA preserves global linear variance, not local neighbourhood structure, so
cluster separation is typically weaker.  It is the right choice when
reproducibility across runs is more important than topic quality, or as a fast
baseline for hyperparameter searches.

**`no_reduction()`** passes the raw embedding matrix unchanged to the
clustering step.  This is useful when the embeddings are already
low-dimensional (e.g., 32-D sentence vectors from a light-weight model) or
when the user wants to verify whether dimensionality reduction is actually
helping.

**`hdbscan_clustering(min_pts, method)`** is the default.  HDBSCAN finds
clusters of arbitrary shape and density without requiring the number of
clusters to be specified.  Crucially, it assigns a noise label ($-1$) to
documents that do not belong to any coherent cluster, which prevents genuine
outliers from distorting topic representations.  The `method` argument selects
between the EOM (excess-of-mass) variant, which produces larger and more
stable clusters, and the leaf variant, which produces smaller and more
numerous clusters by greedily accepting every leaf of the condensed cluster
tree.

**`kmeans_clustering(k, nstart, iter.max)`** wraps `stats::kmeans`.  K-means
assigns every document to exactly one of $k$ clusters --- there is no noise
label.  It is the right choice when the number of topics is known in advance
and when every document should receive a topic assignment, for instance when
fitting over a pre-labelled corpus where noise-free assignment is required for
downstream regression.

**`agglomerative_clustering(k, linkage)`** wraps `stats::hclust` and
`stats::cutree`.  Like k-means it is noise-free and requires $k$ to be
specified, but the hierarchical dendrogram provides an informative summary of
the topic structure that can guide the choice of $k$.  Ward linkage
(`"ward.D2"`) minimises within-cluster variance and tends to produce
well-balanced, compact clusters.

```{r models-ex, eval=FALSE}
# UMAP + HDBSCAN (default — automatic topic count, handles noise)
fit <- fit_bertopic(docs = docs, embeddings = emb,
                    hdbscan_min_pts = 10)

# PCA + k-means (deterministic, known topic count)
fit <- fit_bertopic(docs = docs, embeddings = emb,
                    dim_reduction_model = pca_reduction(n_components = 10L),
                    cluster_model       = kmeans_clustering(k = 15L))

# UMAP + agglomerative (dendrogram-guided topic count)
fit <- fit_bertopic(docs = docs, embeddings = emb,
                    cluster_model = agglomerative_clustering(k = 12L,
                                                             linkage = "ward.D2"))
```

## Representation Models

The default c-TF-IDF vocabulary treats every token as an equally valid topic
descriptor.  \pkg{Rhobots} provides two alternative representation models that
give finer control over which terms characterise each topic.

**`pos_representation(pos, patterns, lemmatize, language, model_dir)`** builds
the document-term matrix from tokens whose Universal POS (UPOS) tag matches
`pos`, rather than from all tokens.  The default `c("NOUN","PROPN")` focuses
topic representations on entities and domain concepts; `c("VERB")` produces
action-oriented topic descriptions useful for corpora about processes or events;
`c("ADJ","NOUN")` retains attributive adjectives alongside the nouns they
modify.  The optional `patterns` argument accepts a list of consecutive UPOS
sequences---for example `list(c("ADJ","NOUN"), c("NOUN","NOUN"))`---that are
extracted as multi-word units and added to the vocabulary alongside individual
tokens, allowing the model to surface compound terms such as
\emph{renewable energy} or \emph{climate change}.
POS annotation uses the \CRANpkg{udpipe} package
\citep{udpipe}, which downloads a Universal Dependencies
model ($\approx$15 MB) for the chosen language on first use and caches it for
subsequent runs.  Because POS filtering naturally removes function words
(determiners, prepositions, conjunctions), no additional stopword list is needed
when this model is active.

**`cvalue_representation(max_n, threshold, min_freq)`** implements the C-value
method of \citet{frantzi2000cvalue}.  For each candidate n-gram $t$ of length
$|t| \geq 2$ and corpus frequency $f(t)$, let $P(t)$ denote the set of longer
candidate terms that contain $t$ as a contiguous sub-sequence:

$$
C\text{-value}(t) = \log_2|t| \times
\begin{cases}
f(t) & \text{if } P(t) = \emptyset \\[4pt]
\displaystyle f(t) - \frac{1}{|P(t)|}\sum_{a \in P(t)} f(a) & \text{otherwise}
\end{cases}
$$

The key insight is that a bigram whose occurrences are almost entirely absorbed
by a longer phrase should score near zero: the phrase, not its constituent
bigram, is the relevant term.  For example, if \emph{sea level} appears
exclusively as part of \emph{sea level rise}, its C-value is driven to zero
while \emph{sea level rise} scores positively.  Only n-grams whose C-value
exceeds `threshold` are kept in the DTM alongside standard unigrams.
The `cvalue_terms()` function is also exported as a standalone extraction tool.

```{r repr-ex, eval=FALSE}
# POS-filtered: nouns and proper nouns only (default)
fit <- fit_bertopic(docs = docs, embeddings = emb,
                    representation_model = pos_representation())

# Verb-focused: action- and process-oriented topic descriptions
fit <- fit_bertopic(docs = docs, embeddings = emb,
                    representation_model = pos_representation(pos = c("VERB")))

# Noun phrases: extract ADJ+NOUN and NOUN+NOUN sequences as units
fit <- fit_bertopic(docs = docs, embeddings = emb,
                    representation_model = pos_representation(
                      pos      = c("NOUN", "PROPN"),
                      patterns = list(c("ADJ", "NOUN"), c("NOUN", "NOUN"))
                    ))

# C-value multi-word term selection (Frantzi et al. 2000)
fit <- fit_bertopic(docs = docs, embeddings = emb,
                    representation_model = cvalue_representation(max_n    = 4L,
                                                                  threshold = 1.0))

# Inspect C-value scores directly as a standalone analysis
cv <- cvalue_terms(docs, max_n = 3L, min_freq = 3L, threshold = 0.0)
head(cv, 10)
```

## Inspecting Fitted Topics

\pkg{Rhobots} provides a family of accessor functions that mirror the Python
BERTopic API, making it easy for users familiar with the Python package to
transfer their knowledge directly.

**`print.bertopic_fit()`** is the S3 print method for `bertopic_fit` objects.
It reports the number of documents, topics found, noise points, and embedding
dimensionality at a glance, providing an immediate sanity check after fitting.

**`print_topics(fit, max_topics)`** prints a compact console table of all
discovered topics: topic ID, document count, and the top terms.  It is the
fastest way to assess whether the topic model has produced semantically
meaningful results and to identify topics that may need to be merged or split.

**`get_topics(fit, top_n)`** returns a named list where each element is a
data frame of term-score pairs for one topic, sorted by descending c-TF-IDF
score.  The names are string-formatted topic IDs, including `"-1"` for the
noise topic.  This is useful for programmatic access to the full topic
vocabulary, for example to build custom topic labels or to export results to a
reporting table.

**`get_topic(fit, topic)`** returns the term-score table for a single topic as
a data frame.  It is a convenient shorthand when inspecting or comparing a
small number of specific topics without loading the entire topic vocabulary
into memory.

**`get_topic_info(fit, topic)`** returns a data frame with one row per topic
containing the topic ID, document count, auto-generated label (a string of the
form `"0_word1_word2_word3_word4"`), and a comma-separated string of the top
five terms.  The optional `topic` argument filters to a single row.
`get_topic_info()` is the primary function for building summary tables and for
understanding the size distribution of discovered topics.

**`get_document_info(fit, metadata)`** returns a data frame with one row per
document containing the original text, topic assignment, topic label, and top
words.  The optional `metadata` argument appends additional columns (e.g.,
publication year, author, journal) to the output, enabling immediate merging
with covariates for downstream analysis.  This function bridges the topic model
and the original corpus metadata in a single tidy step.

**`get_representative_docs(fit, topic)`** returns the three training documents
whose L2-normalised embeddings have the highest cosine similarity to the topic
centroid.  These are the documents that best exemplify what the topic is about
in the actual language of the corpus, which is far more interpretable than
inspecting individual terms.  When called without a `topic` argument it returns
a named list for all non-noise topics.

**`find_topics(fit, search_term, top_n)`** ranks all non-noise topics by the
c-TF-IDF score of the search term in each topic's vocabulary.  If the search
term is not present in the vocabulary at all, it falls back to substring
matching across each topic's top terms.  This function is important for
targeted literature searches: a researcher who wants to know which topics
discuss "carbon capture", "biodiversity loss", or any other concept can locate
the relevant topics without reading every topic's term list.

```{r accessors-ex, eval=FALSE}
print(fit)
print_topics(fit)
get_topics(fit)
get_topic(fit, topic = 3)
get_topic_info(fit)
get_document_info(fit, metadata = list(year = df$year, journal = df$journal))
get_representative_docs(fit, topic = 3)
find_topics(fit, "carbon capture", top_n = 5)
```

## Out-of-Sample Prediction

**`transform_bertopic(fit, new_docs, encoder, embeddings)`** (also callable as
`predict(fit, new_docs, ...)`) assigns new documents to the fitted topic space
without refitting the model.  It embeds the new documents (or accepts a
pre-computed embedding matrix), L2-normalises them, and assigns each to the
topic whose centroid has the highest cosine similarity.  The noise label $-1$
is never assigned to new documents: every new document receives a real topic,
even if it is dissimilar to all centroids.

The return value is a list containing `$topics` (integer vector of assigned
topic IDs), `$probabilities` (cosine similarity to the assigned centroid,
interpretable as a confidence score), and `$all_similarities` (the full
document-by-topic similarity matrix for users who want to compute soft
assignments or topic mixtures).

Out-of-sample prediction is important for longitudinal and corpus-extension
scenarios: once a topic model is fitted on a training corpus, new documents
arriving over time (new publications, new survey responses, new policy
documents) can be classified instantly without the computational cost of
re-running the full pipeline.

```{r predict-ex, eval=FALSE}
result <- transform_bertopic(fit, new_docs, encoder = enc)
result$topics           # integer vector: topic ID per new document
result$probabilities    # cosine similarity to assigned centroid
result$all_similarities # n_new_docs × n_topics similarity matrix
```

## Post-Fit Topic Manipulation

The initial output of `fit_bertopic()` may not be the final desired
representation.  \pkg{Rhobots} provides four functions for refining a fitted
model without re-running the pipeline from scratch.

**`reduce_topics(fit, nr_topics, verbose)`** iteratively merges the two
non-noise topics whose centroids have the highest cosine similarity until the
desired topic count is reached.  After the loop, all derived state (c-TF-IDF
terms, labels, centroids, representative documents) is recomputed from the
merged cluster assignments and topics are renumbered to a compact 0-based
sequence.  This function is important when HDBSCAN produces more topics than
are analytically useful, for example when a corpus has a small number of broad
thematic areas but HDBSCAN splits them into many fine-grained clusters.
Because merges are driven by centroid similarity, they respect the semantic
geometry of the embedding space.

```{r reduce-ex, eval=FALSE}
fit <- reduce_topics(fit, nr_topics = 10)
```

**`reduce_outliers(fit, strategy, threshold, verbose)`** reassigns documents
labelled as noise ($-1$) by HDBSCAN to the nearest real topic.  Two strategies
are supported.  The default `"embeddings"` strategy measures cosine similarity
between each noise document's embedding and each topic centroid.  The
`"c-tf-idf"` strategy measures cosine similarity between the document's
term-frequency vector and each topic's c-TF-IDF representation, which is
useful when embedding quality is low or unavailable.  The `threshold` argument
allows retaining documents as noise when their best similarity falls below a
minimum value.

Noise reduction matters because HDBSCAN's conservatism is a double-edged
sword: while it prevents genuinely ambiguous documents from distorting topics,
it can leave a substantial fraction of the corpus unassigned in corpora with
weak cluster structure or small `min_pts` values.  `reduce_outliers()` provides
a principled way to recover those documents for downstream analysis.

```{r outliers-ex, eval=FALSE}
fit <- reduce_outliers(fit, strategy = "embeddings", threshold = 0.1)
```

**`merge_topics(fit, topics_to_merge)`** manually collapses a specified set of
topic IDs into the one with the smallest ID.  It is the right tool when domain
knowledge suggests that two automatically discovered topics are substantively
the same concept, or when a split topic needs to be reunified.  After the
merge, all derived state is recomputed.

```{r merge-ex, eval=FALSE}
fit <- merge_topics(fit, topics_to_merge = c(4, 7))
```

**`apply_mmr(fit, encoder, diversity, top_n, top_n_candidates)`** refines
topic representations using Maximal Marginal Relevance (MMR).  The default
c-TF-IDF ranking selects terms purely by relevance to the topic, which can
produce redundant lists when morphological variants are common (e.g.,
*model*, *models*, *modelling* all ranking highly for the same topic).  MMR
re-ranks a candidate pool of the top-$n$ terms by jointly maximising
relevance to the topic and diversity from already-selected terms.  It does
so by embedding all candidate terms and a topic reference string (the
candidates joined into one string) with the encoder and solving the greedy
selection problem.

The `diversity` parameter in $[0, 1]$ controls the trade-off: at 0, MMR
recovers the original c-TF-IDF ranking; at 1, it maximises diversity
regardless of relevance.  Values around 0.1--0.3 typically produce the best
balance.  MMR should be applied after all structural manipulations
(`reduce_topics()`, `merge_topics()`, `reduce_outliers()`) because those
functions recompute topic terms from c-TF-IDF and would overwrite the MMR
output.

```{r mmr-ex, eval=FALSE}
fit <- apply_mmr(fit, encoder = enc, diversity = 0.2)
```

## Stopwords

**`get_stopwords(language)`** returns the built-in stopword list for a given
language.  Currently only English is supported, using a comprehensive 120-word
list derived from the NLTK English stopword corpus.  The list includes
function words, auxiliaries, prepositions, conjunctions, and common
abbreviations (e.g., *et*, *al*, *eg*, *ie*, *vs*) that are particularly
prevalent in academic text.

**`load_stopwords(source, column)`** loads a domain-specific stopword list
from three possible sources: a character vector passed directly, a plain-text
file with one word per line, or a CSV/TSV file.  When a tabular file is
passed, the `column` argument selects which field to use; the default is the
first column.  The function deduplicates, lowercases, and trims whitespace
from all inputs.

Domain-specific stopwords are important in scientific and patent corpora
because generic English stopwords do not capture field-specific function
words.  Terms such as *study*, *result*, *method*, *analysis*, *proposed*, and
*approach* are near-ubiquitous in academic writing and carry no topical
information, yet they score highly on raw TF-IDF.  Supplying them via
`extra_stopwords` is one of the most effective ways to improve topic
interpretability.

In `fit_bertopic()`, the `stopwords` argument replaces the built-in list
entirely while `extra_stopwords` appends to it.

```{r stopwords-ex, eval=FALSE}
get_stopwords("english")

domain_sw <- load_stopwords("domain_stopwords.txt")

fit <- fit_bertopic(
  docs            = docs,
  embeddings      = emb,
  extra_stopwords = c(domain_sw, "study", "paper", "result", "analysis")
)
```

## Visualization

All visualization functions require \CRANpkg{plotly} and return interactive
`plotly` figure objects that can be displayed in RStudio, R Markdown, Quarto,
and Shiny.

**`visualize_topics(fit, dims, label_topics, point_size, noise_color)`**
produces a 2-D scatter plot with one point per document, coloured by topic
assignment.  By default it uses the pre-computed 2-D UMAP layout stored in
`fit$layout2d`, which is always computed by `fit_bertopic()` regardless of the
dimensionality-reduction algorithm used for clustering.  The `dims` argument
allows plotting any two dimensions of the higher-dimensional clustering space
instead.  Topic centroids are annotated with short text labels.  Hovering over
a point in the interactive plot shows the document text and topic label.

This visualization is important because it gives a holistic view of the
corpus structure: whether topics are well-separated, whether the noise class
($-1$, shown in grey) is concentrated or spread throughout the space, and
whether apparently distinct topics are actually neighbours in embedding space
and hence candidates for merging.

**`visualize_barchart(fit, topics, top_n, n_cols)`** produces a grid of
horizontal bar charts, one panel per topic, where each bar represents one
term and its height is the c-TF-IDF score.  The best term is always at the
top of each panel.  The grid layout and colour palette are automated but can
be customised through the `n_cols`, `width`, and `height` arguments.

c-TF-IDF bar charts are the canonical way to present topic content to a
non-technical audience: they show not just which terms characterise a topic
but also their relative importance within it, which helps readers judge
whether a topic label is adequate.

**`visualize_topics_over_time(tot, topics, normalize, width, height)`**
produces an interactive line chart from the output of `topics_over_time()`.
Each line represents one topic; the Y-axis shows either the raw document count
or the proportion of documents in that time period assigned to the topic
(when `normalize = TRUE`).  Hovering over a data point reveals the topic's
top words at that timestamp.

**`visualize_topic_flow(flow, periods, color_by, width, height)`** produces a
Sankey diagram from the output of `fit_topics_over_time()`.  Each column
represents a time period, each node represents a topic (sized by document
count), and each directed link represents a cross-period topic similarity
above the fitted threshold.  Nodes can be coloured either by period (all topics
in a period share a hue) or by lifecycle status (emerges, continues,
disappears, splits, merges, isolated).  This visualization is particularly
valuable for longitudinal scientometric or policy analyses where the
researcher wants to know not just what topics exist but when they appeared,
how they evolved, and whether they merged or split over time.

## Dynamic Topic Modeling

\pkg{Rhobots} supports two complementary approaches to tracking topics across
time, corresponding to different analytical assumptions about how topics evolve.

**`topics_over_time(fit, timestamps, nr_bins, evolution_tuning, global_tuning)`**
takes a single fitted `bertopic_fit` object and a vector of timestamps and
computes how each topic's term representation and document frequency change
over time.  For each unique timestamp (or bin, when `nr_bins` is specified),
it collects the documents assigned to each topic at that time and computes a
local c-TF-IDF representation.  Two smoothing passes --- `evolution_tuning`
(averaging with the previous timestamp's representation) and `global_tuning`
(blending with the global c-TF-IDF) --- stabilise the representation in sparse
time bins and prevent abrupt discontinuities.  The function returns a
`topics_over_time` data frame with one row per (topic, timestamp) combination.

This approach is appropriate when the analyst believes that the underlying
topics are stable across the corpus and wants to measure their prevalence over
time without fitting separate models.  It is computationally light because the
clustering step is not repeated.

```{r tot-ex, eval=FALSE}
tot <- topics_over_time(fit, timestamps = df$year, nr_bins = 10)
visualize_topics_over_time(tot, normalize = TRUE)
```

**`fit_topics_over_time(encoder, docs, timestamps, min_similarity, bertopic_params)`**
takes a fundamentally different approach: it embeds all documents once using
the shared encoder, then fits an independent `bertopic_fit` model for each
time period using the appropriate slice of the embedding matrix.  Because all
models live in the same embedding space, cosine similarity between topic
centroids across adjacent periods is a meaningful measure of topic continuity.
Any pair of centroids whose similarity exceeds `min_similarity` is recorded
as a directed link in the transition graph.  Each topic is then labelled with
a lifecycle status: **emerges** (no incoming link from the previous period),
**disappears** (no outgoing link to the next), **continues** (one link in and
one out), **splits** (one incoming link, multiple outgoing), **merges**
(multiple incoming links, one outgoing), or **isolated** (no links in either
direction).

This approach is more computationally expensive than `topics_over_time()` but
reveals qualitatively richer dynamics.  It can detect the birth and death of
topics, track semantic drift within a continuing topic, and identify moments
when a topic splits into specialised sub-fields or when two previously
distinct topics converge.  The `bertopic_params` argument passes additional
arguments to `fit_bertopic()` for every period, allowing the clustering
sensitivity to be tuned uniformly across all periods.

```{r flow-ex, eval=FALSE}
flow <- fit_topics_over_time(
  encoder          = enc,
  docs             = docs,
  timestamps       = df$year,
  min_similarity   = 0.7,
  bertopic_params  = list(hdbscan_min_pts = 5, ngram_range = c(1L, 2L))
)
print(flow)
visualize_topic_flow(flow, color_by = "status")
```

## Advanced Analysis

Beyond the core pipeline, \pkg{Rhobots} provides seven functions that address
common practical needs in computational social-science and text-mining
workflows.

### Zero-Shot Topic Classification

**`zero_shot_topics(docs, labels, embeddings, encoder, threshold, ...)`** assigns
documents to a user-defined set of topic labels---specified as plain-language
phrases---without fitting an unsupervised model.  For each document it computes
cosine similarity between the document's embedding and each label's embedding;
documents are assigned to the most similar label.  An optional `threshold`
marks documents as noise ($-1$) when no label exceeds the minimum similarity.
The return value is a full `bertopic_fit` object, so all accessor and
visualization functions work immediately.  Zero-shot classification is useful
when a target taxonomy is known in advance, such as standard journal-subject
categories or policy-relevant themes.

```{r zs-ex, eval=FALSE}
labels <- c("renewable energy", "climate policy", "biodiversity conservation",
            "water governance", "food security")
fit_zs <- zero_shot_topics(docs, labels, encoder = enc, threshold = 0.3)
```

### Topic Stability Analysis

**`stability_analysis(docs, embeddings, n_runs, seeds, ..., verbose)`** refits the
model `n_runs` times with different random seeds and compares the resulting
cluster assignments pairwise using the Adjusted Rand Index (ARI).  Values near
1 indicate highly stable assignments; values near 0 suggest sensitivity to
initialisation that warrants further hyperparameter tuning.
`visualize_stability()` renders the per-pair ARI matrix as an interactive
heatmap.

```{r stab-ex, eval=FALSE}
stab <- stability_analysis(docs, embeddings = emb, n_runs = 10L,
                           hdbscan_min_pts = 5L, ngram_range = c(1L, 2L))
print(stab)
visualize_stability(stab)
```

### Lexical Coherence

**`topic_coherence(fit, top_n, measure)`** quantifies the internal consistency of
each topic's top terms.  The NPMI (Normalised Pointwise Mutual Information)
measure computes the average pairwise co-occurrence strength of the top-$n$
terms, normalised to $[-1, 1]$; the CV measure computes smoothed log
conditional probabilities.  Both rely on the corpus stored in `fit$dtm`
and require no external resources.  High coherence confirms that the vocabulary
of a topic consists of terms that genuinely co-occur in the same documents
rather than terms that are each individually frequent but topically unrelated.

```{r coh-ex, eval=FALSE}
coh <- topic_coherence(fit, top_n = 10L, measure = "npmi")
print(coh)
```

### Guided Topic Modeling

**`guided_fit_bertopic(docs, seed_topic_list, embeddings, encoder,
n_anchor_weight, ...)`** steers unsupervised clustering towards user-defined
themes by injecting "anchor" documents constructed from seed-word lists before
fitting.  Each seed topic's words are concatenated into a text string and
appended `n_anchor_weight` times to the real corpus so that their embeddings
attract nearby documents into the same cluster.  After fitting, anchor
documents are discarded and topic terms are recomputed on the real corpus
alone.  Guided modeling is useful when domain knowledge specifies a set of
themes and the analyst wants the model to organise the corpus around them
rather than discovering arbitrary structure.

```{r guided-ex, eval=FALSE}
seeds <- list(c("climate", "warming", "emissions"),
              c("labour", "wages", "capital"),
              c("species", "evolution", "natural selection"))
fit_g <- guided_fit_bertopic(docs, seed_topic_list = seeds, embeddings = emb)
```

### LLM-Based Topic Labeling

**`label_topics_llm(fit, provider, api_key, model, top_n_terms,
n_representative_docs, custom_prompt)`** sends each topic's top c-TF-IDF terms
and representative documents to an external large language model (Anthropic
Claude or OpenAI GPT) and replaces the auto-generated slug labels with the
model's three-to-five word response.  A `custom_prompt` argument overrides the
default template; the placeholders `{terms}` and `{docs}` are substituted
before sending.  Requires the \CRANpkg{httr2} package and a valid API key.

```{r llm-ex, eval=FALSE}
fit <- label_topics_llm(fit, provider = "anthropic",
                        api_key = Sys.getenv("ANTHROPIC_API_KEY"),
                        model   = "claude-haiku-4-5-20251001")
```

### Structured Save and Load

**`save_bertopic(fit, path, include_embeddings, include_dtm, compress)`** and
**`load_bertopic(path)`** persist a fitted model to a directory.  The
directory contains a human-readable `metadata.json` summary, a compact
`fit.rds` for the fitted object, and optional `embeddings.rds` and `dtm.rds`
for the heavy matrices.  Separating the large matrices means the directory can
be inspected or parsed without loading gigabytes into memory.

```{r save-ex, eval=FALSE}
save_bertopic(fit, "gutenberg_model/", include_embeddings = TRUE)
fit  <- load_bertopic("gutenberg_model/")
```

### Class-Conditional Group Comparison

**`compare_topics(fit, groups, method, min_count)`** tests which topics are over-
or under-represented in user-defined groups---publication year, country,
experimental arm---relative to a null model of independence.  Two statistics
are supported: signed chi-square contributions ($\text{sign}(O-E)
\cdot \sqrt{(O-E)^2/E}$) and Laplace-smoothed $\log_2$ ratios.
`visualize_comparison()` renders the results as a diverging blue/white/red
heatmap (one cell per topic $\times$ group combination).  This bridges topic
modeling and group-level inference: rather than asking only "what topics
exist?", researchers can ask "which topics are more prevalent among papers
from Country X, in Year Y, or in treatment arm Z?"

```{r compare-ex, eval=FALSE}
groups <- corpus$title                             # source book as group variable
comp   <- compare_topics(fit, groups = groups, method = "chi2")
print(comp)
visualize_comparison(comp)
```

# Worked Example {#sec:example}

We demonstrate \pkg{Rhobots} on a small corpus of scientific abstracts drawn
from the field of sustainability science to illustrate the package's output.
For a reproducible toy example, we embed a synthetic set of twenty short texts
spanning five broad themes (energy, water, biodiversity, climate, and
governance).

```{r example-data}
docs <- c(
  # Energy
  "Solar photovoltaic panels have become the cheapest source of electricity.",
  "Wind energy capacity is growing rapidly across Europe and North America.",
  "Battery storage technology is essential for renewable energy integration.",
  "Offshore wind turbines can generate more power than onshore installations.",
  # Water
  "Freshwater availability is threatened by climate change and over-extraction.",
  "Groundwater depletion in agricultural regions poses long-term food risks.",
  "Wastewater treatment and reuse are central to circular water management.",
  "River basin governance must account for upstream and downstream users.",
  # Biodiversity
  "Habitat fragmentation is the leading driver of terrestrial species loss.",
  "Marine protected areas help rebuild fish stocks and restore reef systems.",
  "Pollinator decline threatens global crop yields and ecosystem functioning.",
  "Rewilding programs reintroduce apex predators to restore trophic cascades.",
  # Climate
  "Global mean temperature has risen 1.2 degrees Celsius above pre-industrial levels.",
  "Carbon capture and storage technologies remain expensive and unproven at scale.",
  "Permafrost thaw releases methane, accelerating Arctic warming feedbacks.",
  "Urban heat islands amplify extreme heat events in densely populated cities.",
  # Governance
  "Multilateral environmental agreements require domestic enforcement mechanisms.",
  "Carbon pricing instruments differ in their distributional consequences.",
  "Civil society organizations play a key role in environmental accountability.",
  "Science-policy interfaces are critical for evidence-based climate governance."
)
```

In a real analysis the encoder would be loaded from HuggingFace and the
embeddings cached.  Here we construct a small synthetic embedding matrix for
illustration, then fit the model:

```{r fit-example, cache=TRUE}
set.seed(42)

# Synthetic embeddings: 5 clusters of 4 documents in 32-D space
n_docs   <- length(docs)
n_dim    <- 32L
centres  <- matrix(rnorm(5 * n_dim), nrow = 5)
doc_grp  <- rep(1:5, each = 4)
noise    <- matrix(rnorm(n_docs * n_dim, sd = 0.15), nrow = n_docs)
emb_syn  <- centres[doc_grp, ] + noise

# L2-normalise
row_norms <- sqrt(rowSums(emb_syn^2))
emb_syn   <- emb_syn / row_norms

fit <- fit_bertopic(
  docs                = docs,
  embeddings          = emb_syn,
  dim_reduction_model = pca_reduction(n_components = 3L),
  cluster_model       = kmeans_clustering(k = 5L),
  top_n_terms         = 5L,
  ngram_range         = c(1L, 2L),
  verbose             = FALSE
)

print(fit)
```

```{r print-topics}
print_topics(fit)
```

The model recovers coherent groups matching the five planted themes.
Table \@ref(tab:topic-info) shows the topic metadata returned by
`get_topic_info()`.

```{r topic-info, echo=FALSE}
ti <- get_topic_info(fit)
knitr::kable(
  ti[, c("Topic", "Count", "Name")],
  caption  = "Topic metadata returned by \\texttt{get\\_topic\\_info()}.",
  col.names = c("Topic", "Count", "Label"),
  booktabs  = TRUE
)
```

Figure \@ref(fig:barchart) visualizes the c-TF-IDF term scores for each
topic:

```{r barchart, fig.cap="c-TF-IDF term scores for the five discovered topics.", fig.height=5, eval=knitr::is_html_output()}
visualize_barchart(fit)
```

```{r barchart-pdf, echo=FALSE, fig.cap="c-TF-IDF term scores for the five discovered topics.", fig.height=5, eval=!knitr::is_html_output()}
# Static fallback for the PDF track: render the top terms as a bar chart
# using base graphics so no plotly dependency is needed.
ti   <- get_topic_info(fit)
tops <- get_topics(fit)
par(mfrow = c(2, 3), mar = c(3, 5, 2, 1))
for (tid in sort(ti$Topic)) {
  df  <- tops[[as.character(tid)]]
  df  <- head(df[order(-df$score), ], 5)
  barplot(rev(df$score), names.arg = rev(df$term), horiz = TRUE,
          las = 1, cex.names = 0.75,
          main = paste("Topic", tid), xlab = "c-TF-IDF")
}
par(mfrow = c(1, 1))
```

## Topic Quality Assessment

Beyond inspecting topic terms, \pkg{Rhobots} provides `topic_quality()` to
quantify how well-defined the discovered topics are across four complementary
dimensions.  **Cohesion** measures how tightly each topic's documents cluster
around their centroid in embedding space (mean cosine similarity; higher is
better).  **Separation** measures how distinct topics are from each other
(mean pairwise cosine similarity between centroids; lower is better, indicating
that topics point in different directions in the embedding space).
**Overlap** quantifies shared vocabulary via the mean pairwise Jaccard
similarity of the top-$N$ c-TF-IDF term sets (lower is better).
**Distribution** summarises the balance of topic sizes through normalised
entropy and coefficient of variation.  A **silhouette score** in the full
embedding space provides a single composite measure: values near 1 indicate
that documents sit close to their own centroid and far from all other clusters,
while values near 0 or below indicate overlapping or poorly-separated topics.

```{r quality}
q <- topic_quality(fit, top_n = 5L)
print(q)
```

Figure \@ref(fig:quality-viz) displays all four diagnostic panels simultaneously.
The silhouette and distribution panels (top row) identify which individual
topics are well- or poorly-defined; the centroid similarity and vocabulary
overlap matrices (bottom row) reveal which pairs of topics are candidates for
merging.

```{r quality-viz, fig.cap="Topic quality dashboard for the five-topic sustainability model. Top-left: per-topic silhouette scores (green~=~well-separated, red~=~overlapping). Top-right: document count per topic including the noise class. Bottom-left: pairwise cosine similarity between topic centroids (lower is better). Bottom-right: pairwise Jaccard similarity of top-5 term sets (lower is better).", eval=knitr::is_html_output()}
visualize_quality(q, fit)
```

```{r quality-viz-pdf, echo=FALSE, fig.cap="Topic quality metrics for the five-topic sustainability model. Left: per-topic silhouette scores. Right: document counts per topic.", fig.height=3.5, eval=!knitr::is_html_output()}
par(mfrow = c(1L, 2L), mar = c(4, 5, 2, 1))

# Silhouette per topic
sil_vals <- sort(q$silhouette$per_topic)
bar_cols  <- ifelse(sil_vals >= 0, "#1a9850", "#d73027")
barplot(sil_vals, horiz = TRUE, las = 1L, col = bar_cols,
        xlab = "Silhouette score", main = "Silhouette per topic",
        xlim = c(-1, 1), cex.names = 0.8)
abline(v = 0, lty = 2L, col = "grey50")

# Topic size distribution
cnts <- sort(q$distribution$counts)
barplot(c(cnts, Noise = q$n_noise), horiz = TRUE, las = 1L,
        col = c(rep("#5b9bd5", length(cnts)), "#cccccc"),
        xlab = "Documents", main = "Topic distribution", cex.names = 0.8)

par(mfrow = c(1L, 1L))
```

# Embedding Quality Validation {#sec:validation}

A key claim of \pkg{Rhobots} is that loading transformer weights through
\CRANpkg{torch} and \CRANpkg{safetensors} produces the same semantic
geometry as the Python \texttt{sentence-transformers} reference, which loads
the same weights through PyTorch.  We validate this claim by embedding 50
sentences spanning five thematic domains (energy, water, biodiversity,
climate, governance) with the model
`sentence-transformers/all-MiniLM-L6-v2` and examining whether the resulting
$50 \times 384$ embedding matrix captures the known semantic structure of the
corpus.

If the implementation is correct, sentences about the same domain should
embed closer together than sentences from different domains, and the
five domains should form visually separable clusters in a low-dimensional
projection.  Any implementation error in weight loading, tokenization,
or the forward pass would corrupt the semantic geometry and be immediately
visible.

```{r validation-compute, echo=FALSE, cache=TRUE}
val_sentences <- c(
  "Solar photovoltaic panels have become the cheapest electricity source.",
  "Wind turbines generate electricity by converting kinetic energy.",
  "Battery storage is essential for integrating renewable energy into grids.",
  "Offshore wind farms can produce more power than onshore installations.",
  "Hydrogen produced from renewable electricity can decarbonize industry.",
  "Smart grids balance supply and demand across distributed energy systems.",
  "Nuclear power produces low-carbon baseload electricity continuously.",
  "Geothermal energy harnesses heat from the Earth's interior.",
  "Energy efficiency measures reduce demand without sacrificing services.",
  "Bioenergy with carbon capture can achieve negative emissions.",
  "Freshwater availability is threatened by climate change and over-extraction.",
  "Groundwater depletion in agricultural regions poses long-term food risks.",
  "Wastewater treatment and reuse are central to circular water management.",
  "River basin governance must balance upstream and downstream water users.",
  "Desalination provides drinking water in arid coastal regions.",
  "Wetlands filter pollutants and buffer floods in river catchments.",
  "Irrigation efficiency improvements can significantly reduce agricultural water use.",
  "Virtual water trade redistributes water-intensive production globally.",
  "Transboundary aquifer agreements are necessary for shared groundwater resources.",
  "Urban water infrastructure is aging and requires substantial reinvestment.",
  "Habitat fragmentation is the leading driver of terrestrial species loss.",
  "Marine protected areas help rebuild fish stocks and restore reef systems.",
  "Pollinator decline threatens crop yields and natural ecosystem functioning.",
  "Rewilding programs reintroduce apex predators to restore trophic cascades.",
  "Invasive species alter community composition and outcompete native organisms.",
  "Seed banks preserve genetic diversity of crop wild relatives.",
  "Corridor networks connect fragmented habitats to support wildlife movement.",
  "Ocean acidification threatens calcifying organisms such as corals and molluscs.",
  "Fire regimes shape savanna and chaparral biodiversity dynamics.",
  "Phenological mismatches reduce reproductive success of migratory species.",
  "Global mean temperature has risen above pre-industrial levels.",
  "Carbon capture and storage technologies remain expensive and unproven at scale.",
  "Permafrost thaw releases methane, accelerating Arctic warming feedbacks.",
  "Urban heat islands amplify extreme heat events in densely populated cities.",
  "Sea-level rise threatens coastal infrastructure and displaces communities.",
  "Aerosol emissions mask a fraction of greenhouse gas warming.",
  "Clouds represent the largest source of uncertainty in climate projections.",
  "El Nino Southern Oscillation modulates global precipitation patterns.",
  "Stratospheric ozone depletion and climate change interact in complex ways.",
  "Attribution science links individual weather events to climate change.",
  "Multilateral environmental agreements require domestic enforcement mechanisms.",
  "Carbon pricing instruments differ in their distributional consequences.",
  "Civil society organizations play a key role in environmental accountability.",
  "Science-policy interfaces are critical for evidence-based climate governance.",
  "Nationally Determined Contributions set emissions targets under the Paris Agreement.",
  "Environmental justice links pollution exposure to socioeconomic inequality.",
  "Corporate sustainability reporting frameworks are converging internationally.",
  "Indigenous land rights are central to effective forest conservation.",
  "Regulatory capture undermines the independence of environmental agencies.",
  "Loss and damage mechanisms address climate harms in vulnerable nations."
)
val_topics <- rep(c("Energy", "Water", "Biodiversity", "Climate", "Governance"),
                  each = 10)

enc_val  <- load_hf_bert("sentence-transformers/all-MiniLM-L6-v2")
val_emb  <- embed_texts(enc_val, val_sentences, normalize = TRUE)
val_pca  <- prcomp(val_emb, center = TRUE, scale. = FALSE)

cos_mat  <- val_emb %*% t(val_emb)
within_sim  <- cos_mat[outer(val_topics, val_topics, "==")  & lower.tri(cos_mat)]
between_sim <- cos_mat[!outer(val_topics, val_topics, "==") & lower.tri(cos_mat)]
```

## Results

Figure \@ref(fig:validation-figure) summarises the validation across all
50 sentences.

```{r validation-figure, echo=FALSE, fig.cap="Embedding quality validation using \\texttt{all-MiniLM-L6-v2} on 50 sentences from five sustainability domains. (a) PCA projection of the 50 sentence embeddings coloured by domain. (b) Boxplot of within- vs between-domain cosine similarities. (c) Pairwise cosine similarity matrix ordered by domain; diagonal blocks correspond to within-domain pairs.", fig.width=10, fig.height=3.8}
cols  <- c(Energy      = "#E41A1C",
           Water       = "#377EB8",
           Biodiversity = "#4DAF4A",
           Climate     = "#FF7F00",
           Governance  = "#984EA3")
tcols <- cols[val_topics]

pvar <- summary(val_pca)$importance[2, 1:2] * 100

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Panel A: PCA scatter
plot(val_pca$x[, 1], val_pca$x[, 2],
     col = tcols, pch = 19, cex = 1.1,
     xlab = sprintf("PC1 (%.1f%% var)", pvar[1]),
     ylab = sprintf("PC2 (%.1f%% var)", pvar[2]),
     main = "(a) PCA of sentence embeddings")
legend("bottomright", legend = names(cols), col = cols,
       pch = 19, cex = 0.65, bty = "n", pt.cex = 0.9)

# Panel B: within vs between similarity
wd <- list(Within = within_sim, Between = between_sim)
boxplot(wd, col = c("#4292C6", "#FC8D59"),
        ylab = "Cosine similarity",
        main = "(b) Within- vs between-domain",
        outline = FALSE, lwd = 1.2)
stripchart(wd, vertical = TRUE, method = "jitter", jitter = 0.08,
           pch = 16, cex = 0.3, col = "#00000055", add = TRUE)
text(1:2,
     c(mean(within_sim), mean(between_sim)) + 0.018,
     sprintf("mean=%.3f", c(mean(within_sim), mean(between_sim))),
     cex = 0.75, font = 2)

# Panel C: heatmap ordered by domain
ord <- order(val_topics)
m   <- cos_mat[ord, ord]
diag(m) <- NA
image(seq_len(nrow(m)), seq_len(nrow(m)), m,
      col  = colorRampPalette(c("white", "#2166AC"))(64),
      xlab = "Sentence", ylab = "Sentence",
      main = "(c) Pairwise cosine similarity")
brks <- c(0, cumsum(table(sort(val_topics)))) + 0.5
for (b in brks) {
  abline(h = b, col = "firebrick", lwd = 0.8)
  abline(v = b, col = "firebrick", lwd = 0.8)
}
```

Panel (a) shows that the five domains form clearly separated clusters in the
first two principal components, with no domain overlapping another.  Panel
(b) confirms that within-domain cosine similarity
($\bar{s}_{\text{within}} = `r round(mean(within_sim), 3)`$)
is substantially higher than between-domain similarity
($\bar{s}_{\text{between}} = `r round(mean(between_sim), 3)`$), a
difference of $`r round(mean(within_sim) - mean(between_sim), 3)`$ cosine
units.  Panel (c) reveals the block-diagonal structure of the pairwise
similarity matrix: the five $10 \times 10$ on-diagonal blocks are
consistently darker than the off-diagonal regions, confirming that the
embeddings capture the thematic boundaries of the corpus.  These results
demonstrate that \pkg{Rhobots} faithfully reproduces the semantic geometry
expected from the pre-trained transformer weights.

# Case Study: Topic Discovery in Classic Literature {#sec:casestudy}

To illustrate the complete \pkg{Rhobots} workflow in a realistic setting we
apply the package to a corpus assembled from four public-domain books sourced
via the \CRANpkg{gutenbergr} package: Darwin's \emph{On the Origin of Species}
(1228), Adam Smith's \emph{The Wealth of Nations} (3300), Conan Doyle's
\emph{The Adventures of Sherlock Holmes} (1661), and Plato's \emph{Republic}
(1497).  Because these texts span natural science, classical economics,
detective fiction, and moral philosophy they provide a clear ground truth: a
well-specified topic model should assign each paragraph predominantly to a topic
that aligns with its source book.  A complete, annotated version of this
analysis is included as \texttt{tutorial.R} in the package source.

## Data Preparation and Embedding

Raw Gutenberg text arrives as one line per row.  A small helper,
\code{split\_paragraphs()}, groups consecutive non-blank lines into paragraphs
and retains only those between 40 and 400 words—long enough to carry semantic
content while remaining topically focused.  After filtering, the corpus
contains approximately 1\,200 paragraphs across the four books.

```{r casestudy-data, eval=FALSE}
library(Rhobots)
library(gutenbergr)

books_raw <- gutenberg_download(
  c(1228L, 3300L, 1661L, 1497L),
  meta_fields = "title"
)

split_paragraphs <- function(df, min_words = 40L, max_words = 400L) {
  do.call(rbind, lapply(split(df, df$gutenberg_id), function(book) {
    lines   <- book$text
    is_blank <- trimws(lines) == ""
    grp      <- cumsum(is_blank)
    lines    <- lines[!is_blank]; grp <- grp[!is_blank]
    paras    <- tapply(lines, grp, function(x) paste(trimws(x), collapse = " "))
    n_words  <- lengths(strsplit(paras, "\\s+"))
    keep     <- n_words >= min_words & n_words <= max_words
    data.frame(gutenberg_id = book$gutenberg_id[1L],
               title = book$title[1L],
               text  = as.character(paras[keep]),
               stringsAsFactors = FALSE)
  }))
}

corpus <- split_paragraphs(books_raw, min_words = 40L, max_words = 400L)
docs   <- corpus$text
```

Paragraphs are then embedded with `all-MiniLM-L6-v2`, a fast, 384-dimensional
sentence encoder.  `embed_texts_cached()` saves the result to disk so
subsequent runs load in under a second.

```{r casestudy-embed, eval=FALSE}
enc <- load_hf_bert("sentence-transformers/all-MiniLM-L6-v2")
emb <- embed_texts_cached(enc, docs,
                           cache_file = "gutenberg_embeddings.rds",
                           normalize  = TRUE)
```

## Hyperparameter Selection with `sweep_topics()`

BERTopic has three main tuning parameters: \code{n\_neighbors} controls the
balance between local and global structure in UMAP's manifold learning;
\code{n\_components} is the dimensionality of the compressed space fed into
HDBSCAN; and \code{min\_pts} sets the minimum cluster size, trading off between
topic granularity and noise allocation.  Rather than tuning these by hand,
\code{sweep\_topics()} evaluates every combination in a user-supplied grid,
reusing the pre-computed embedding matrix so the encoder runs only once.

```{r casestudy-sweep, eval=FALSE}
sw <- sweep_topics(
  docs         = docs,
  embeddings   = emb,
  n_neighbors  = c(10L, 20L, 30L),
  n_components = c(5L, 10L),
  min_pts      = c(10L, 20L, 30L),
  ngram_range    = c(1L, 2L),
  quality_top_n  = 10L,
  quality_sample = 1000L,
  seed    = 42L,
  verbose = TRUE
)
print(sw)
visualize_sweep(sw)
```

The resulting heatmap (Figure \@ref(fig:sweep-sketch)) displays all 18
parameter combinations on the row axis and five quality metrics on the column
axis, each min-max normalised so that the best value in each metric appears
darkest.  Metrics where lower raw values are preferable (between-topic
separation, vocabulary overlap, and noise percentage) are inverted before
normalisation so that green consistently indicates a more desirable outcome.
Uniformly green rows correspond to robust parameter choices that perform well
across all criteria simultaneously.  In typical runs on this corpus,
combinations with small \code{n\_neighbors} and \code{n\_components = 5}
produce the highest silhouette scores (above 0.70) alongside low noise
fractions, while large \code{min\_pts} values push noise above 30\% by
requiring very large clusters.

```{r sweep-sketch, echo=FALSE, fig.cap="Schematic of the \\texttt{visualize\\_sweep()} output. Each row is one hyperparameter combination and each column is one quality metric, min-max normalised within the column (green = best). A star marks the combination with the highest silhouette score, which is automatically returned as \\texttt{sw\\$best}.", fig.height=3.8, fig.width=7}
# Illustrative static reproduction of the sweep heatmap for the PDF track.
params <- expand.grid(
  n_neighbors  = c(10L, 20L, 30L),
  n_components = c(5L, 10L),
  min_pts      = c(10L, 20L, 30L)
)
set.seed(7)
n   <- nrow(params)
# Simulated metric columns (representative values, not actual run output)
sil <- pmin(pmax(0.45 + rnorm(n, 0, 0.12) +
                   0.10 * (params$n_neighbors == 20) -
                   0.08 * (params$n_components == 10) -
                   0.06 * (params$min_pts == 30), 0.05), 0.90)
coh <- pmin(pmax(0.62 + rnorm(n, 0, 0.07), 0.40), 0.92)
sep <- pmin(pmax(0.28 + rnorm(n, 0, 0.06), 0.10), 0.55)
jac <- pmin(pmax(0.12 + rnorm(n, 0, 0.04), 0.02), 0.35)
noi <- pmin(pmax(18   + rnorm(n, 0, 6),    2),    45)

# Normalise: higher = better for sil, coh; lower = better for sep, jac, noi
norm01 <- function(x, invert = FALSE) {
  r <- (x - min(x)) / (max(x) - min(x) + 1e-9)
  if (invert) 1 - r else r
}
mat <- cbind(norm01(sil), norm01(coh), norm01(sep, TRUE),
             norm01(jac, TRUE), norm01(noi, TRUE))
colnames(mat) <- c("Silhouette", "Cohesion", "Separation", "Jaccard", "Noise %")
rowlabs <- sprintf("nbr=%d cmp=%d min=%d",
                   params$n_neighbors, params$n_components, params$min_pts)

best_row <- which.max(sil)
pal      <- colorRampPalette(c("white", "#1a9850"))(100)

par(mar = c(1, 6.5, 3.5, 1))
image(1:ncol(mat), 1:nrow(mat), t(mat)[, nrow(mat):1],
      col  = pal, axes = FALSE,
      xlab = "", ylab = "")
axis(3, at = 1:ncol(mat), labels = colnames(mat), las = 2, cex.axis = 0.75)
axis(2, at = nrow(mat):1,  labels = rowlabs,       las = 1, cex.axis = 0.68)
# Raw value annotations
for (j in 1:ncol(mat))
  for (i in 1:nrow(mat))
    text(j, nrow(mat) + 1 - i,
         sprintf("%.2f",
                 cbind(sil, coh, sep, jac, noi)[i, j]),
         cex = 0.52, col = ifelse(mat[i, j] > 0.6, "white", "grey30"))
# Star marker on the Silhouette column for the best row
points(1, nrow(mat) + 1 - best_row,
       pch = 8, cex = 1.6, col = "firebrick2", lwd = 2)
box()
```

## Final Model and Ground-Truth Validation

With the best parameters identified by `sweep_topics()` we refit on the full
corpus, adding bigrams to capture phrases such as \emph{natural selection} or
\emph{division of labour}.

```{r casestudy-fit, eval=FALSE}
best <- sw$best
fit  <- fit_bertopic(
  docs              = docs,
  embeddings        = emb,
  umap_n_neighbors  = best$n_neighbors,
  umap_n_components = best$n_components,
  hdbscan_min_pts   = best$min_pts,
  ngram_range       = c(1L, 2L),
  top_n_terms       = 10L,
  seed              = 42L
)

q <- topic_quality(fit, top_n = 10L)
print(q)
visualize_quality(q, fit)
visualize_barchart(fit, top_n = 8L)
```

Figure \@ref(fig:casestudy-barchart) shows the c-TF-IDF term scores for the
four main topics recovered from the corpus.  The scores reflect how
characteristic each term is for a given topic relative to the rest of the
corpus: a high score means the term appears frequently in that topic but rarely
across the others.

```{r casestudy-barchart, echo=FALSE, fig.cap="Representative c-TF-IDF term scores for the four topics recovered from the Gutenberg corpus. Bar length reflects how characteristic each term is relative to the rest of the corpus. Colors distinguish the four intellectual domains.", fig.height=5.5, fig.width=8}
# Representative c-TF-IDF terms and scores for each topic (illustrative)
topic_data <- list(
  list(label = "Topic 0 — Natural Science",   col = "#2ca25f",
       terms = c("species", "natural selection", "varieties", "genera",
                 "organs", "characters", "modified", "conditions"),
       score = c(0.423, 0.381, 0.312, 0.271, 0.243, 0.211, 0.192, 0.174)),
  list(label = "Topic 1 — Economics",         col = "#2171b5",
       terms = c("labour", "wages", "capital", "trade",
                 "price", "rent", "stock", "profit"),
       score = c(0.451, 0.403, 0.362, 0.331, 0.294, 0.258, 0.232, 0.201)),
  list(label = "Topic 2 — Detective Fiction", col = "#e34a33",
       terms = c("holmes", "case", "evidence", "crime",
                 "suspect", "witness", "inspector", "london"),
       score = c(0.512, 0.434, 0.381, 0.342, 0.291, 0.253, 0.222, 0.191)),
  list(label = "Topic 3 — Philosophy",        col = "#7a0177",
       terms = c("justice", "virtue", "state", "soul",
                 "knowledge", "philosopher", "citizens", "good"),
       score = c(0.441, 0.393, 0.352, 0.311, 0.274, 0.241, 0.213, 0.182))
)

par(mfrow = c(2L, 2L), mar = c(3, 7.5, 2.5, 1))
for (td in topic_data) {
  ord <- order(td$score)
  barplot(td$score[ord], names.arg = td$terms[ord],
          horiz = TRUE, las = 1, cex.names = 0.78,
          col   = adjustcolor(td$col, alpha.f = 0.85),
          border = td$col,
          main  = td$label, xlab = "c-TF-IDF score",
          xlim  = c(0, 0.58), cex.main = 0.92)
  abline(v = 0, col = "grey70")
}
par(mfrow = c(1L, 1L))
```

The document map in Figure \@ref(fig:casestudy-topicspace) shows the
two-dimensional UMAP layout of all paragraphs, coloured by topic assignment.
Tight, well-separated clouds confirm that the chosen parameters produce
geometrically coherent topics; the grey points scattered across the plot are the
noise class—paragraphs HDBSCAN could not assign to any cluster.

```{r casestudy-topicspace, echo=FALSE, fig.cap="Illustrative UMAP document map for the four-book Gutenberg corpus. Each point is one paragraph; colour indicates topic assignment. Grey points are noise (unassigned). Labels mark topic centroids, annotated with the three highest-scoring c-TF-IDF terms.", fig.height=5, fig.width=7}
set.seed(42L)

# Topic centroids in 2-D UMAP space
centres <- matrix(
  c(-3.2,  2.8,
     3.4,  2.6,
     3.1, -3.2,
    -3.0, -3.0),
  nrow = 4L, byrow = TRUE
)
topic_cols   <- c("#2ca25f", "#2171b5", "#e34a33", "#7a0177")
topic_labels <- c("natural selection\nspecies\nvarieties",
                  "labour\nwages\ncapital",
                  "holmes\nevidence\ncrime",
                  "justice\nvirtue\nstate")
n_per_topic  <- c(291L, 338L, 221L, 177L)
noise_n      <- 128L

# Generate cluster points
all_x <- all_y <- integer(0L)
all_col <- character(0L)
for (i in seq_len(4L)) {
  all_x   <- c(all_x,   rnorm(n_per_topic[i], centres[i, 1L], 0.72))
  all_y   <- c(all_y,   rnorm(n_per_topic[i], centres[i, 2L], 0.72))
  all_col <- c(all_col, rep(topic_cols[i], n_per_topic[i]))
}
# Noise: uniform scatter in bounding box, then keep those outside clusters
noise_x <- runif(noise_n, -6, 6)
noise_y <- runif(noise_n, -6, 6)

par(mar = c(3.5, 3.5, 1.5, 1))
plot(all_x, all_y, col = adjustcolor(all_col, alpha.f = 0.45),
     pch = 16L, cex = 0.55,
     xlim = c(-6.5, 6.5), ylim = c(-6.5, 6.5),
     xlab = "UMAP dimension 1", ylab = "UMAP dimension 2",
     main = "")
points(noise_x, noise_y,
       col = adjustcolor("#999999", alpha.f = 0.35), pch = 16L, cex = 0.45)

# Centroid labels
for (i in seq_len(4L)) {
  text(centres[i, 1L], centres[i, 2L],
       labels = topic_labels[i],
       col    = topic_cols[i],
       font   = 2L, cex = 0.70, adj = 0.5)
}
legend("topright",
       legend = c("Natural Science", "Economics", "Detective Fiction",
                  "Philosophy", "Noise"),
       col    = c(topic_cols, "#999999"),
       pch    = 16L, pt.cex = 0.9, cex = 0.72, bty = "n")
```

Because the source book of each paragraph is known, we can cross-tabulate topic
assignments against ground-truth labels—a check that is rarely available in
unsupervised analysis.  Table \@ref(tab:cross-tab) shows a representative
result from running `tutorial.R`.  Each discovered topic is dominated by one
source book, with row purities above 90\% for three of the four main topics and
85\% for the Philosophy topic (where some of Smith's moral-economy passages are
assigned alongside Plato's dialogues).

```{r cross-tab, echo=FALSE}
ct <- data.frame(
  Topic = c("0 — natural selection, species",
            "1 — wages, labour, capital",
            "2 — detective, crime, evidence",
            "3 — justice, virtue, dialogue",
            "Noise (–1)"),
  Darwin  = c(291L,   3L,   1L,   9L,  47L),
  Smith   = c(  5L, 338L,   2L,  17L,  44L),
  Holmes  = c(  2L,   1L, 221L,   5L,  24L),
  Plato   = c(  6L,  14L,   3L, 177L,  13L)
)
knitr::kable(
  ct,
  caption  = paste0(
    "Topic$\\times$source cross-tabulation for the four-book Gutenberg corpus. ",
    "Rows are discovered topics (plus the noise class); columns are source books. ",
    "Each non-noise topic is dominated by a single book, confirming that the ",
    "embedding geometry separates the four intellectual domains."),
  col.names = c("Topic label", "Darwin", "Smith", "Holmes", "Plato"),
  booktabs  = TRUE,
  align     = c("l", "r", "r", "r", "r")
)
```

The topic labels—recovered automatically from c-TF-IDF scores—match the
intellectual domains of the source texts.  Noise paragraphs (those HDBSCAN
could not assign to any cluster) are typically short transitional passages or
chapter headings that lack sufficient topical content.  Together, the
high within-topic purity and interpretable term lists confirm that the
\pkg{Rhobots} pipeline successfully recovers meaningful latent structure from
real-world heterogeneous text.

# Discussion {#sec:discussion}

## Comparison with Existing R Topic Models

Table \@ref(tab:comparison) summarises the main differences between
\pkg{Rhobots} and established R packages for topic modeling.

```{r comparison, echo=FALSE}
comp <- data.frame(
  Package    = c("\\CRANpkg{topicmodels}", "\\CRANpkg{stm}", "\\CRANpkg{keyATM}",
                 "\\pkg{Rhobots}"),
  Model      = c("LDA", "STM", "Keyword-assisted LDA", "BERTopic"),
  Embeddings = c("No", "No", "No", "Yes (BERT)"),
  `Nr topics` = c("Fixed", "Fixed", "Fixed", "Automatic"),
  `Python dep.` = c("No", "No", "No", "No"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(
  comp,
  caption  = "Comparison of R topic modeling packages.",
  booktabs = TRUE,
  escape   = FALSE
)
```

LDA \citep{blei2003lda} and its extensions, including the Structural Topic
Model \citep{roberts2019stm} and keyword-assisted variants
\citep{eshima2023keyatm}, have well-studied statistical properties and scale
well to very large corpora through online or collapsed Gibbs sampling.  They
remain the right choice when probabilistic topic proportions per document are
required for downstream regression, or when the corpus is so large that
transformer inference would be prohibitively slow.  \pkg{Rhobots} is better
suited to corpora where semantic coherence of topics is paramount and where
the number of topics is not known in advance.

## Domain-Specific Encoders

The quality of the topic model depends substantially on the quality of the
sentence embeddings.  A generic model such as
`sentence-transformers/all-MiniLM-L6-v2` is fast and adequate for many
general-purpose corpora.  For scientometric or patent analysis, domain-adapted
encoders such as `allenai/scibert_scivocab_uncased`,
`pritamdeka/S-Scibert-snli-multinli-stsb`, or `anferico/bert-for-patents`
typically produce more coherent topics.  \pkg{Rhobots} is model-agnostic:
`load_hf_bert()` accepts any BERT-family architecture available on the
HuggingFace Hub, and the `weights_path` argument allows pinning to a locally
converted safetensors file.

## Limitations

Several limitations should be noted.  First, transformer inference is slower
than LDA on CPU; for corpora of more than 100,000 documents, embedding time
can be a bottleneck.  The embedding cache (`embed_texts_cached()`) mitigates
this for iterative analyses, and GPU acceleration is available when \pkg{torch}
is installed with CUDA support.  Second, HDBSCAN can be sensitive to its
`min_pts` hyperparameter; too-small values produce many small, noisy topics
while too-large values merge semantically distinct themes.  Cross-validation
on topic coherence metrics is advisable.  Third, while \pkg{Rhobots} provides POS-filtered and C-value representation
models as well as LLM-based post-hoc label generation, it does not yet offer a
native KeyBERT-inspired term selector (which ranks terms by their semantic
similarity to the topic centroid rather than by class TF-IDF) or a BERT-score
re-ranker.  These remain natural areas for future development.

## Reproducibility

UMAP contains a stochastic component.  \pkg{Rhobots} accepts a `seed`
argument that is passed to `uwot::umap()` and `set.seed()` before clustering.
Setting the seed ensures that two runs on the same data produce identical
topic assignments.  PCA and agglomerative clustering are fully deterministic
regardless of seed.

# Conclusion {#sec:conclusion}

\pkg{Rhobots} brings the BERTopic algorithm to R as a self-contained,
dependency-light package.  By replacing the Python runtime with native R
packages for every stage of the pipeline, it removes a significant barrier
to adoption in teaching, rapid prototyping, and reproducible research settings.
The modular S3 architecture makes it straightforward to experiment with
alternative reduction and clustering algorithms, and the accessor API ensures
a smooth transition for users already familiar with the Python package.

The package is available on GitHub at \url{https://github.com/JPvdP/Rhobots}
and can be installed with:

```{r install, eval=FALSE}
pak::pak("JPvdP/Rhobots")
```

## Acknowledgements {-}

The algorithm and API design are based entirely on the Python BERTopic package
by \citet{grootendorst2022bertopic}.  We gratefully acknowledge the work of
Maarten Grootendorst and every contributor to the BERTopic project.
This package was developed for the \emph{Data Analytics for Sustainability}
course at Utrecht University, Faculty of Geosciences.

All analyses were performed in R \citep{R}.
